---
title: "CFB Data Plays Examples"
description: "Using the CFB Data Plays Endpoint to pull down the 2020 season by week using <code>cfbd_plays()</code>"
author: "Saiem Gilani <br><a href='https://twitter.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    creator: "@saiemgilani"
    card: summary_large_image
    site: "@cfbfastR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Load and Install Packages**

```{r cfbd_plays, eval=FALSE}
if (!requireNamespace('pacman', quietly = TRUE)){
  install.packages('pacman')
}
pacman::p_load(dplyr,tidyr, cfbfastR)
#pacman::p_load_current_gh("sportsdataverse/cfbfastR")
```

### **Pull first 3 weeks of 2020 season using `cfbd_plays()`**

```{r cfbd_plays_purrr, eval = FALSE}
year_vector <- 2020
week_vector <- 1:3
weekly_year_df <- expand.grid(year = year_vector, week = week_vector)

tictoc::tic()
year_split <- split(weekly_year_df, weekly_year_df$year)
for (i in 1:length(year_split)) {
  progressr::with_progress({
    year_split[[i]] <- year_split[[i]] |>
      dplyr::mutate(
        pbp = purrr::map2(
          .x = year,
          .y = week,
          cfbd_plays,
          season_type = "both"
        )
      )

    Sys.sleep(1)
  })
}

tictoc::toc()

year_split <- lapply(year_split, function(x) {
  x |> tidyr::unnest(pbp, names_repair = "minimal")
})

all_years <- dplyr::bind_rows(year_split)
glimpse(all_years)
```

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/JaredDLee' target='blank'><img src='https://img.shields.io/github/followers/JaredDLee?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@JaredDLee'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/arbitanalytics' target='blank'><img src='https://img.shields.io/github/followers/arbitanalytics?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@arbitanalytics'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTex Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani, Akshay Easwaran, Jared Lee, Eric Hess},
  title = {cfbfastR: The SportsDataverse's R Package for College Football Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2026}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package
