---
title: "CFB Data Teams Examples"
description: "Get team rosters (<code>cfbd_team_rosters()</code>), talent (<code>cfbd_team_talent()</code>) and team matchup history (<code>cfbd_team_matchup_history()</code>) and records (<code>cfbd_team_matchup_records()</code>) from the CFBD API"
author: "Saiem Gilani, Akshay Easwaran, Jared Lee <br><a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a> <a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    creator: "@saiemgilani"
    card: summary_large_image
    site: "@cfbfastR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Load and Install Packages**

```{r}
if (!requireNamespace('pak', quietly = TRUE)){
  install.packages('pak')
}
pak::pak(c("dplyr", "tidyr", "gt", "cfbfastR"))
library(dplyr)
library(tidyr)
library(gt)
library(cfbfastR)
# pak::pak("sportsdataverse/cfbfastR")
```

### **Get Team Info**

```{r team_info_ex}
cfbd_team_info(conference = "SEC")

cfbd_team_info(conference = "Ind")

cfbd_team_info(year = 2019)
```

### **Get Team Matchup History (Total Record)**

```{r team_matchup_records_ex}

cfbd_team_matchup_records("Texas", "Oklahoma")

cfbd_team_matchup_records("Texas A&M", "TCU", min_year = 1975)

```

### **Get Team Matchup History**

```{r team_matchup_ex}
cfbd_team_matchup("Texas", "Oklahoma")

cfbd_team_matchup("Texas A&M", "TCU")

cfbd_team_matchup("Texas A&M", "TCU", min_year = 1975)

cfbd_team_matchup("Florida State", "Florida", min_year = 1975)
```

### **Get Team Rosters**

```{r team_roster_ex}

cfbd_team_roster(year = 2013, team = "Florida State")

```

### **Get Team Talent**

```{r team_talent_ex}
cfbd_team_talent()

cfbd_team_talent(year = 2018)

```

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/Kazink36' target='blank'><img src='https://img.shields.io/github/followers/Kazink36?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@Kazink36'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/ehess' target='blank'><img src='https://img.shields.io/github/followers/ehess?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@ehess'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
<a href='https://github.com/michaelegle' target='blank'><img src='https://img.shields.io/github/followers/michaelegle?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@michaelegle'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
<a href='https://github.com/natemanzo' target='blank'><img src='https://img.shields.io/github/followers/natemanzo?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@natemanzo'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
<a href='https://github.com/CFBNumbers' target='blank'><img src='https://img.shields.io/github/followers/CFBNumbers?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
<a href='https://github.com/tejseth' target='blank'><img src='https://img.shields.io/github/followers/tejseth?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@tejseth'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
<a href='https://github.com/mcqconor' target='blank'><img src='https://img.shields.io/github/followers/mcqconor?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@mcqconor'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
<a href='https://github.com/tanho63' target='blank'><img src='https://img.shields.io/github/followers/tanho63?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@tanho63'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
<a href='https://github.com/keegan-abdoo' target='blank'><img src='https://img.shields.io/github/followers/keegan-abdoo?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@keegan-abdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
<a href='https://github.com/Maatspencer' target='blank'><img src='https://img.shields.io/github/followers/Maatspencer?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
<a href='https://github.com/mrcaseb' target='blank'><img src='https://img.shields.io/github/followers/mrcaseb?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
<a href='https://github.com/john-b-edwards' target='blank'><img src='https://img.shields.io/github/followers/john-b-edwards?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@john-b-edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>
<a href='https://github.com/bradisbrad' target='blank'><img src='https://img.shields.io/github/followers/bradisbrad?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@bradisbrad'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTeX Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani and Akshay Easwaran and Jared Lee and Eric Hess},
  title = {cfbfastR: Access College Football Play by Play Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2021}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package
