---
title: "CFB Data Stats Examples"
description: "Settling <strong>2019 LSU</strong> and <strong>2013 Florida State</strong> offense debates using Team and Player Stats from the CFBD API"
author: "Saiem Gilani <br><a href='https://twitter.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    creator: "@saiemgilani"
    card: summary_large_image
    site: "@cfbfastR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Load and Install Packages**

```{r}
if (!requireNamespace('pak', quietly = TRUE)){
  install.packages('pak')
}
pak::pak(c("dplyr", "tidyr", "gt", "cfbfastR"))
library(dplyr)
library(tidyr)
library(gt)
library(cfbfastR)
# pak::pak("sportsdataverse/cfbfastR")
```

## Settling **2019 LSU** and **2013 Florida State** offense debates

### **Get Season Statistics by Team**

```{r team_season_stats}
team_season_stats <- dplyr::bind_rows(
   cfbd_stats_season_team(year=2019, team = "LSU"),
   cfbd_stats_season_team(year=2013, team = "Florida State")
)
logos <- read.csv("https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/logos.csv")
logos<- logos |> dplyr::select(-"conference")
df_team_season <- team_season_stats |>
   dplyr::left_join(logos, by=c("team"="school"))
```

```{r team_season_long}
df_team_season_long <- as.data.frame(t(as.matrix(df_team_season)))
colnames(df_team_season_long) <- df_team_season$team
```

### **Get Season Advanced Statistics by Team**

```{r team_season_adv}
df_team_season_adv <- dplyr::bind_rows(
   cfbd_stats_season_advanced(2019, team = "LSU"),
   cfbd_stats_season_advanced(2013, team = "Florida State")
)
df_team_season_adv <- df_team_season_adv |>
   dplyr::left_join(logos, by=c("team"="school"))

```

### **Get Game Advanced Stats**

```{r team_game_adv}
df_team_game_adv <- dplyr::bind_rows(
   cfbd_stats_game_advanced(2019, team = "LSU"),
   cfbd_stats_game_advanced(2013, team = "Florida State")
)
df_team_game_adv <- df_team_game_adv |>
   dplyr::left_join(logos, by=c("team"="school"))
```

### **Get Season Statistics by Player**

```{r passing_stats_table_ex, fig.alt="A table showing passing statistics for LSU in 2019 and Florida State in 2013" }
source("https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/gt_theme_code_SG.R")
passing_df <- dplyr::bind_rows(
   cfbd_stats_season_player(2019, team = "LSU", category = "passing"),
   cfbd_stats_season_player(2013, team = "Florida State", category = "passing")) |>
   dplyr::left_join(logos, by=c("team"="school")) |>
   dplyr::group_by(team) |>
   dplyr::mutate(logo = sprintf('<img src="%s" height="30" alt="%s logo">', logo, team)) |>
   dplyr::select(logo,
                 player,
                 passing_completions,
                 passing_att,
                 passing_yds,
                 passing_td,
                 passing_int,
                 passing_ypa) |>
   arrange( desc(passing_yds), team)
passing_df |> gt() |>
  tab_header(title = "Passing Summary") |>
  cols_label(logo="",
             player = "Player",
             passing_completions = "C",
             passing_att = "Att",
             passing_yds = "Yds",
             passing_td = "TDs",
             passing_int = "INTs",
             passing_ypa = "YPA") |>
  data_color(
    columns = c("passing_yds"),
    colors = scales::col_numeric(
      palette = "RdBu",
      domain = c(-6000,6000)
    )
  ) |>
  data_color(
    columns = c("passing_td"),
    colors = scales::col_numeric(
      palette = "RdBu",
      domain = c(-60,60)
    )
  ) |>
  data_color(
    columns = c("passing_td"),
    colors = scales::col_numeric(
      palette = "RdBu",
      domain = c(-60,60)
    )
  ) |>
  # Render the team logos from pre-built <img> HTML that carries alt text
  # (gt::web_image() cannot set alt; important for screen-reader accessibility).
  fmt_markdown(columns = "logo") |>
  tab_source_note(source_note = md("**Table:** @SaiemGilani | **Data:** @CFB_Data with @cfbfastR v2.0.0")) |>
  gt_theme_538(table.width = px(550))

```

### **College Football Mapping for Stats Categories**

```{r}

cfbd_stats_categories()

```

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/JaredDLee' target='blank'><img src='https://img.shields.io/github/followers/JaredDLee?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@JaredDLee'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/arbitanalytics' target='blank'><img src='https://img.shields.io/github/followers/arbitanalytics?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@arbitanalytics'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTex Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani, Akshay Easwaran, Jared Lee, Eric Hess},
  title = {cfbfastR: The SportsDataverse's R Package for College Football Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2026}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package
