---
title: "ESPN College Football Cookbook"
description: "A recipe-driven tour of the <code>espn_cfb_*()</code> family in <code>cfbfastR</code> -- teams, games, play-by-play, modeled EPA/WPA, and ratings."
author: "Saiem Gilani <br><a href='https://twitter.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
# The ESPN endpoints are live web APIs. We only want this vignette to *run*
# its code chunks when there is a working connection -- on CRAN, on an
# offline machine, or behind a firewall, the chunks should still render
# (showing the code) but skip execution so the build never fails.
has_espn <- FALSE
try({
  con <- url("https://site.api.espn.com", open = "rb")
  close(con)
  has_espn <- TRUE
}, silent = TRUE)

knitr::opts_chunk$set(
  echo = TRUE,
  eval = has_espn,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

This is a *cookbook*. Instead of marching through every function in the
`espn_cfb_*()` family one at a time, we are going to answer a handful of real
questions -- the kind you actually ask when you sit down with college football
data -- and pick up the functions, and the *shape* of the function names, as we
go.

```{r load-pkg}
library(cfbfastR)
library(dplyr)
```

### A note on the function names

Before the first recipe, one idea worth planting early, because it pays off in
every recipe after it. The ESPN layer of `cfbfastR` is named on a strict,
predictable pattern:

```
espn_cfb_<entity>_<detail>
```

`<entity>` is the *thing* you are asking about -- a `team`, a `game`, a
`player`. `<detail>` is the *slice* of that thing you want -- a `schedule`, a
`record`, a `roster`. The plural form (`espn_cfb_teams()`) is the catalog of
every entity; the singular form (`espn_cfb_team()`) is one entity in depth.

The practical upshot: **once you know one function, you can guess its
siblings.** If `espn_cfb_game_teams()` gives you the two teams in a game, then
`espn_cfb_game_team_roster()` almost certainly gives you a roster for a team in
that game -- and it does. You rarely need to look anything up. You just need to
know what you want and spell it the way the package spells it.

We will use four anchor IDs throughout: game `401628339`, team `61` (Georgia
Tech), athlete `4427191`, and season `2024`.

---

## Recipe 1 -- A team's season at a glance

**The question:** I want a quick scouting snapshot of one program -- who they
are, and how their season went.

Start at the catalog. `espn_cfb_teams()` (plural -- the catalog) is the lookup
table for the whole sport: one row per team, with the all-important `team_id`
you feed everything else.

```{r r1-teams}
teams <- espn_cfb_teams()
teams |>
  filter(team_id == "61") |>
  select(team_id, display_name, abbreviation, location, color)
```

Now zoom in. `espn_cfb_team()` (singular -- one entity, in depth) takes that
`team_id` and a season and returns the team's detail record, including its home
venue.

```{r r1-team}
gt_team <- espn_cfb_team(team_id = 61, year = 2024)
gt_team |>
  select(team_id, display_name, venue_name, venue_city, venue_state)
```

We have the *entity*. Now we want a *detail* of it -- the schedule. The name
writes itself: `espn_cfb_team_schedule()`.

```{r r1-schedule}
gt_sched <- espn_cfb_team_schedule(team_id = 61, year = 2024)
gt_sched |>
  select(any_of(c("game_id", "week", "game_date",
                  "home_team", "away_team"))) |>
  head(6)
```

And the won/lost summary of that schedule -- another detail of the same entity,
so another `espn_cfb_team_*()` function: `espn_cfb_team_record()`.

```{r r1-record}
espn_cfb_team_record(team_id = 61, year = 2024) |>
  select(any_of(c("team_id", "type", "summary", "wins", "losses")))
```

Notice the rhythm of this recipe. We never left the `espn_cfb_team*` shelf:
catalog (`teams`), then entity (`team`), then two details of that entity
(`team_schedule`, `team_record`). That shelf has more on it -- `team_roster()`,
`team_stats()`, `team_coaches()`, `team_leaders()`, `team_events()` -- and you
can now reach for any of them without checking the docs first.

---

## Recipe 2 -- Building a game box score

**The question:** Give me the box score for a single game -- the team totals
and the individual player lines.

We are switching entities, from `team` to `game`, so the prefix switches with
us: `espn_cfb_team_*()` becomes `espn_cfb_game_*()`. Same grammar, new shelf.

First, who played, and what was the score? `espn_cfb_game_teams()` is the
two-team summary of a game. It takes a `format` argument -- `"long"` gives one
row per team (tidy for plotting), `"wide"` gives one row for the whole game with
`home_*` / `away_*` columns (handy for a box-score header).

```{r r2-game-teams}
game_hdr <- espn_cfb_game_teams(game_id = 401628339, format = "wide")
game_hdr |>
  select(any_of(c("game_id", "home_team", "home_team_score",
                  "away_team", "away_team_score")))
```

Now the team statistical totals. We want a *detail* (`statistics`) of a *team*
within a *game* -- read that right to left and the name falls out:
`espn_cfb_game_team_statistics()`.

```{r r2-team-stats}
team_stats <- espn_cfb_game_team_statistics(game_id = 401628339)
team_stats |>
  select(any_of(c("team_id", "team", "stat_name",
                  "display_value"))) |>
  head(10)
```

Finally the individual lines -- the player box score. The `<entity>` is still
`game`, the `<detail>` is the `player_box`:

```{r r2-player-box}
player_box <- espn_cfb_game_player_box(game_id = 401628339)
player_box |>
  select(any_of(c("team", "athlete_display_name", "category",
                  "stat_name", "stat_value"))) |>
  head(10)
```

Three functions, one game, one consistent prefix. If you later want the per-team
leaders, you already know it is `espn_cfb_game_team_leaders()`; if you want each
team's roster *as it appeared in this game*, it is
`espn_cfb_game_team_roster()`. You are guessing the names correctly now because
the names are not really being guessed -- they are being *spelled*.

---

## Recipe 3 -- Play-by-play and who was on the field

**The question:** Walk me through the game play by play -- and tell me which
players were involved in each play.

`espn_cfb_game_pbp()` is the play-by-play detail of a game: one tidy row per
play. On its own that is a clean play log. The interesting part is the
`participants` argument, which decides whether the *people* involved in each
play come along for the ride.

* `participants = "none"` (default) -- just the plays.
* `participants = "wide"` -- the involved athletes spread into columns on the
  play row (`participant_1_*`, `participant_2_*`, ...). Best when you want one
  flat table.
* `participants = "long"` -- one extra row per athlete per play.

```{r r3-pbp}
pbp <- espn_cfb_game_pbp(game_id = 401628339, participants = "wide")
pbp |>
  select(any_of(c("play_id", "period_number", "type_text",
                  "text"))) |>
  head(8)
```

```{r r3-participants}
# The participant_* columns ride alongside each play in "wide" mode.
pbp |>
  select(any_of(grep("participant", names(pbp), value = TRUE))) |>
  head(5)
```

By default these wrappers also enrich every team-id column with readable team
detail -- that is the `team_detail = TRUE` argument you will see on most
`espn_cfb_game_*()` functions. It is what turns a bare `team_id` into a
`team`, `team_abbreviation`, and team colors without a manual join. If you ever
want the raw, un-joined frame -- for a smaller object, or to do the join
yourself -- pass `team_detail = FALSE`.

```{r r3-team-detail}
# team_detail = TRUE is the default; this just makes it explicit.
pbp_plain <- espn_cfb_game_pbp(game_id = 401628339, team_detail = FALSE)
ncol(pbp) - ncol(pbp_plain)  # the team-detail columns that the join adds
```

---

## Recipe 4 -- From drives to plays

**The question:** I do not want a flat play log -- I want the game organized
into *drives*, and then I want to drill from a drive down to its plays.

`espn_cfb_game_drives()` is the drive-level detail of a game: one row per
possession, with the result, yardage, and clock for each. Its `plays` argument
controls how much detail rides along:

* `plays = "none"` (default) -- just the drive summaries.
* `plays = "list"` -- a `plays` list-column, one nested play table per drive.
* `plays = "expand"` -- the drives unnested all the way down to one row per
  play, drive context carried along on every play row (`drive_*` columns).

```{r r4-drives}
drives <- espn_cfb_game_drives(game_id = 401628339)
drives |>
  select(any_of(c("drive_id", "team", "drive_result",
                  "yards", "offensive_plays"))) |>
  head(6)
```

When you want plays *and* their drive context in one flat table, `plays =
"expand"` does it in a single call:

```{r r4-expand}
drive_plays <- espn_cfb_game_drives(game_id = 401628339, plays = "expand")
drive_plays |>
  select(any_of(c("drive_id", "drive_result", "play_id",
                  "type_text"))) |>
  head(8)
```

There is also a two-step route, and it is worth knowing because it shows the
package's split between *fetching* and *transforming*. Pull the drives once with
`plays = "list"`, then flatten with `espn_cfb_unnest_plays()` -- a pure
transform, no second web request:

```{r r4-unnest}
drives_listed <- espn_cfb_game_drives(game_id = 401628339, plays = "list")
unnested <- espn_cfb_unnest_plays(drives_listed)
identical(
  sort(names(unnested)),
  sort(names(drive_plays))
)
```

Same flat table, two roads to it: the `plays = "expand"` shortcut, or
`plays = "list"` plus `espn_cfb_unnest_plays()`. Reach for the second when you
already have a `"list"`-shaped drives object in hand and do not want to pay for
the request again.

---

## Recipe 5 -- Modeled play-by-play with EPA and WPA

**The question:** The play log is fine, but I want the *valuation* -- Expected
Points Added and Win Probability Added on every play.

This is where `cfbfastR` stops being a thin API client and starts being an
analytics package. `espn_cfb_pbp_v2()` is the modern, core-v2-sourced
play-by-play function. With its default `epa_wpa = FALSE` it returns the
assembled play-by-play frame -- structurally the same idea as Recipe 3, just
sourced from the more robust core-v2 drives endpoint.

The payoff is `epa_wpa = TRUE`. Flip that switch and the same EPA/WPA modeling
stack the package has always run -- the `mgcv` GAMs, the expected-points model,
the field-goal and win-probability models -- gets applied to the play frame.

```{r r5-pbp-v2}
pbp_v2 <- espn_cfb_pbp_v2(game_id = 401628339, epa_wpa = TRUE)
pbp_v2 |>
  select(any_of(c("play_id", "down", "distance", "yards_gained",
                  "ep_before", "ep_after", "EPA",
                  "wp_before", "wp_after", "wpa"))) |>
  head(10)
```

`ep_before` / `ep_after` and `EPA` are the expected-points columns;
`wp_before` / `wp_after` and `wpa` are the win-probability ones. From here a
team's offensive EPA per play for the game is a one-liner:

```{r r5-epa-summary}
pbp_v2 |>
  filter(!is.na(EPA), !is.na(pos_team)) |>
  group_by(pos_team) |>
  summarise(plays = dplyr::n(),
            epa_per_play = round(mean(EPA), 3),
            .groups = "drop")
```

The `_v2` suffix is the one place the naming pattern carries a *version* rather
than an entity/detail -- it marks the modern successor to the legacy
`espn_cfb_pbp()`. For new work, prefer `espn_cfb_pbp_v2()`.

---

## Recipe 6 -- QBR, power ratings, and recruiting

**The question:** Step back from a single game -- I want season-level context.
How good is each team, who are the best quarterbacks, and who is signing the
best classes?

These three live on the *ratings* and *catalog* shelves, and they are all keyed
by `year` rather than `game_id` -- a tell that they describe a whole season.

ESPN's quarterback rating, season-wide, is `espn_cfb_qbr()`:

```{r r6-qbr}
qbr <- espn_cfb_qbr(year = 2024)
qbr |>
  select(any_of(c("athlete_display_name", "team_short_name",
                  "qbr_total", "qb_plays"))) |>
  head(8)
```

ESPN's Football Power Index -- their team strength rating -- is
`espn_cfb_powerindex()`:

```{r r6-powerindex}
fpi <- espn_cfb_powerindex(year = 2024)
fpi |>
  filter(stat_name == "fpi") |>
  select(any_of(c("team_id", "display_name", "value"))) |>
  arrange(desc(value)) |>
  head(8)
```

And the recruiting class -- the incoming talent -- is `espn_cfb_recruits()`:

```{r r6-recruits}
recruits <- espn_cfb_recruits(year = 2024, max_results = 25)
recruits |>
  select(any_of(c("rank", "name", "position", "grade",
                  "school_name"))) |>
  head(10)
```

Three different questions -- quarterback quality, team strength, recruiting --
three functions, and every one of them asked for by `year`. When a function
takes a `year` and no `game_id`, you are looking at a season-level summary;
when it takes a `game_id`, you are inside a single game. That single signature
cue tells you the *grain* of the data before you have read a word of the help
page.

---

## A note on caching

Several `espn_cfb_*()` wrappers enrich their output with two slow-changing
catalogs -- the full team list (`espn_cfb_teams()`) and the position list
(`espn_cfb_positions()`). Re-fetching those on every call would be wasteful, so
`cfbfastR` memoises them.

You control the cache backend with the `cfbfastR.cache` option, set **before**
loading the package:

* `"memory"` (default) -- in-memory cache, gone when the R session ends.
* `"filesystem"` -- persistent on-disk cache, survives between sessions.
* `"off"` -- no memoisation; every catalog lookup hits ESPN.

```{r cache-option, eval=FALSE}
# Set this before library(cfbfastR) to persist the catalog cache to disk.
options(cfbfastR.cache = "filesystem")
```

When you want to force a fresh pull of the catalogs -- after a long-running
session, or while debugging -- clear the memoised lookups with
`espn_cfb_clear_cache()`:

```{r cache-clear}
espn_cfb_clear_cache()
```

It returns invisibly and is a safe no-op when caching is `"off"`.

---

## A note on proxies

If you're working from a network that routes outbound HTTP through a corporate
proxy, you don't need to thread a `proxy = ...` argument through every
`cfbd_*()` call. `cfbfastR`'s internal HTTP helper (`get_req()`) resolves a
proxy in this order:

1. An explicit `proxy` argument passed to the wrapper (highest precedence).
2. `getOption("cfbfastR.proxy")` -- a session-level fallback.
3. The standard `http_proxy` / `https_proxy` / `no_proxy` environment variables, which libcurl reads automatically.

The recommended pattern is to set the option once at the top of your script and
let every subsequent call pick it up:

```{r proxy-option, eval=FALSE}
options(cfbfastR.proxy = "http://proxy.host.example:8080")

# Every cfbd_*() and espn_cfb_*() call now routes through the proxy.
teams <- espn_cfb_teams()
plays <- cfbd_pbp_data(year = 2024, week = 1, team = "Georgia")
```

For an authenticated proxy, pass a named list with `url` / `port` / `username` /
`password` / `auth` -- the list is spread directly into `httr2::req_proxy()`:

```{r proxy-auth, eval=FALSE}
options(cfbfastR.proxy = list(
  url      = "http://proxy.host.example",
  port     = 8080,
  username = "me",
  password = "pw",
  auth     = "basic"
))
```

If you prefer the environment-variable path -- handy in CI containers and Docker
images where the proxy is already exported -- nothing extra is needed in R, but
you can also set them from the session:

```{r proxy-env, eval=FALSE}
Sys.setenv(https_proxy = "http://proxy.host.example:8080")
```

The ESPN wrappers (`espn_cfb_*()`, `espn_metrics_*()`, `espn_ratings_*()`) don't
expose a per-call `proxy =` argument, but they go through the same `httr2`
stack and honour the same env-var-based proxy handling, so a single `options()`
or env-var setup covers the whole package.

---

## Where to go next

We covered seven recipes and touched maybe fifteen functions -- but the
`espn_cfb_*()` family has roughly sixty. The point of this cookbook was never
to enumerate them. It was to make the *next* one predictable.

You now know the grammar: `espn_cfb_<entity>_<detail>`, plural for the catalog
and singular for one entity in depth; `team_*` for programs and `game_*` for
single games; a `year` argument means season-level and a `game_id` argument
means inside-a-game. Want each game's win-probability chart inputs? Reach for
`espn_cfb_game_probabilities()`. Want a player's game-by-game log? That is
`espn_cfb_player_gamelog()`. Want the season's weekly rankings? Try
`espn_cfb_week_rankings()`.

You will be right more often than not -- because in this package, guessing the
name and knowing the name are very nearly the same thing.

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/JaredDLee' target='blank'><img src='https://img.shields.io/github/followers/JaredDLee?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@JaredDLee'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/arbitanalytics' target='blank'><img src='https://img.shields.io/github/followers/arbitanalytics?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@arbitanalytics'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTex Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani, Akshay Easwaran, Jared Lee, Eric Hess},
  title = {cfbfastR: The SportsDataverse's R Package for College Football Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2026}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/JaredDLee' target='blank'><img src='https://img.shields.io/github/followers/JaredDLee?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@JaredDLee'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/arbitanalytics' target='blank'><img src='https://img.shields.io/github/followers/arbitanalytics?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@arbitanalytics'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTex Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani, Akshay Easwaran, Jared Lee, Eric Hess},
  title = {cfbfastR: The SportsDataverse's R Package for College Football Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2026}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package
