---
title: "CPOE, xpass, Field Goals and Fourth Downs: The Derived Models (Part VI)"
description: "The six models cfbfastR builds on top of expected points and win probability: completion probability and CPOE, expected pass rate, field goals, two-point conversions, fourth-down decisions and QBR — with their real outputs and their limits."
author: "Saiem Gilani, Akshay Easwaran, Jared Lee"
date: "2026-09-03"
opengraph:
  image:
    src: "https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/social_card_cfbfastR_final_quote.png"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
old <- options(rmarkdown.html_vignette.check_title = FALSE)
```

[Expected points](college-football-expected-points-model-fundamentals-part-i.html)
and [win probability](college-football-expected-points-model-fundamentals-part-v.html)
are the foundation. Everything in this article is built on top of them, and all
of it ships in the same
[`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
bundle.

There is a useful distinction to draw first. Some of these are **models** — they
predict a thing and you read the number. Others are **analytic surfaces**: they
build the game state that *would* follow a choice, score it with the WP model,
and compare. Fourth down and the two-point decision are surfaces, and they are
ports of [`cfb4th`](https://github.com/sportsdataverse/cfb4th).

## The bundle at a glance

| Artifact | Objective | Features | Produces |
|---|---|---|---|
| `ep_model` | `multi:softprob` (7) | 8 | `ep_before`, `ep_after`, `EPA` |
| `wp_naive` | `binary:logistic` | 12 | `wp_before`, `wpa` |
| `wp_spread` | `binary:logistic` | 13 | `vegas_wp` |
| `cfb_cp_model` | `binary:logistic` | 8 | `cp`, `cpoe` |
| `xpass_model` | `binary:logistic` | 7 | `xpass`, `pass_oe` |
| `fg_model` | `binary:logistic` | 5 | field-goal probability |
| `two_pt_model` | `binary:logistic` | 4 | `prob_2pt` |
| `fd_model` | `multi:softprob` (**76**) | 9 | fourth-down gain distribution |
| `qbr_model` | `reg:squarederror` | 10 | QBR |

Plus `punt_distribution.parquet`, which is not a model at all — see below.

## Completion probability and CPOE

`cp` is the probability a given pass attempt is completed, from down, distance,
field position, score, clock, home field, period and a passing-down indicator.
**CPOE** is the difference between what happened and what was expected, on a
percentage-point scale:

$$\mathrm{CPOE} = 100 \times (\text{completion} - \mathrm{cp})$$

| Situation | `cp` |
|---|---|
| 1st & 10, own 25 | 0.639 |
| 3rd & 3, opponent 40 | 0.595 |
| 2nd & 20, own 20 | 0.627 |
| 3rd & 12, own 30 | 0.489 |

CPOE is the standard way to separate a quarterback's accuracy from his
situation. A 65% completion rate throwing 3rd-and-12s is a much better
performance than 65% throwing 1st-and-10s, and CPOE is what says so.

Note that the model has **no receiver, pressure, or air-yards input** — it is a
situational baseline, not a throw-difficulty model. A quarterback checking down
constantly will post good CPOE.

## Expected pass rate (`xpass` and `pass_oe`)

`xpass` is the probability a play *is* a pass, given the situation — a model of
tendency, not of quality. `pass_oe` (pass rate over expected) is
`100 * (pass - xpass)`, and measures how much more pass-happy a team is than its
situations warrant.

| Situation | `xpass` |
|---|---|
| 1st & goal, opponent 3 | 0.141 |
| 2nd & 2, own 33 | 0.247 |
| 3rd & 2, opponent 40 | 0.255 |
| 1st & 10, own 25 | 0.373 |
| 2nd & 8, own 27 | 0.536 |
| 3rd & 8, opponent 40 | 0.782 |

The spread is the point: 14% on 1st and goal from the 3, 78% on 3rd and 8. Any
raw pass-rate statistic is mostly measuring which of these situations a team
found itself in. `pass_oe` removes that.

### An era-encoding trap

The `xpass` model uses an **ordinal** `era` feature cutting at **2006 / 2013 /
2017**. The FG, two-point and QBR models use **one-hot** `era0`–`era3` cutting at
**2006 / 2013 / 2020**.

Different encodings *and* different cutpoints, in the same bundle. If you score
these models by hand, build each one's era feature from its own rule — reusing
one for the other is silent and wrong. `cfbfastR` handles this internally; it is
listed here because it is exactly the kind of thing that survives a code review
and fails in production.

## The field goal model

Five features: `yards_to_goal` and the one-hot era. Kick distance is
`yards_to_goal + 17`.

| `yards_to_goal` | Kick | P(make) |
|---|---|---|
| 2 | 19 | 0.956 |
| 10 | 27 | 0.893 |
| 17 | 34 | 0.809 |
| 20 | 37 | 0.754 |
| 25 | 42 | 0.675 |
| 30 | 47 | 0.614 |
| 33 | 50 | 0.542 |
| 38 | 55 | 0.455 |
| 43 | 60 | 0.366 |
| 48 | 65 | 0.366 |

A 50-yarder is a coin flip, which is a useful number to carry around.

**Note the tail.** The curve is flat at 0.366 from 60 yards out — the model
returns the same probability for a 60-yard and a 65-yard attempt. There is
almost no training data out there (nobody attempts 65-yarders except in
desperation), so the model has nothing to distinguish them and extrapolates a
constant. Treat anything beyond about 57 yards as "the model has run out of
evidence" rather than a real estimate.

The model also has **no kicker identity, weather or altitude input**. A kick in
Laramie and a kick in a dome get the same number.

## The two-point model, and an honest caveat

Four features: `posteam_spread`, `posteam_total`, `pos_score_diff`, `era`. It
produces `prob_2pt`, which feeds the two-point decision surface —
`two_pt_wp`, `xp_wp`, `two_pt_wp_diff` and `two_pt_recommendation`.

Scoring it in an even game with a 55-point total returns **0.578, and does not
move with the score margin**:

| Score margin | `prob_2pt` |
|---|---|
| −2 | 0.578 |
| −1 | 0.578 |
| 0 | 0.578 |
| +1 | 0.578 |
| +2 | 0.578 |

Two things follow, and both are documented rather than hidden.

First, **0.578 is optimistic.** Real college two-point conversion rates run
closer to 45%. The `NEWS` entry says as much: the bundled model is optimistic
about college conversion rates.

Second, because the recommendation compares win probabilities computed from that
number, **read `two_pt_wp_diff` rather than `two_pt_recommendation`.** `cfb4th`'s
rule has no margin — it will recommend going for two on a difference of 0.0001 —
and with an optimistic conversion probability that tips too often toward the
aggressive call. The magnitude tells you whether the decision is close; the bare
recommendation does not.

The surface itself is sound: it was verified **bit-identical to
`sportsdataverse-py` to eight decimal places**. The caveat is about the input,
not the arithmetic.

## The fourth-down decision

This is the most elaborate surface in the package, and the 76-class model is the
interesting part.

`fd_model` is a `multi:softprob` classifier over **76 classes: yards gained, from
−10 to +65**. It does not predict whether a fourth down is converted. It predicts
the *whole distribution* of what happens if you go for it.

The go branch then:

1. Scores the 76-class distribution for the fourth-down state.
2. Expands the play into **one hypothetical game state per outcome** — every
   yardage result gets its own down, distance and field position.
3. Scores each with the win probability model.
4. Averages, weighted by the distribution.

That average is `go_wp` — the win probability of going for it, integrated over
everything that could happen.

The punt branch is different again. **`cfb4th` does not model punts.** It joins
an empirical distribution of punt end-yardlines from
`punt_distribution.parquet`, expands to one state per landing spot, scores each,
and averages. The field-goal branch uses the FG model above, weighting a make
against a miss and the field position each produces.

The outputs:

| Column | Meaning |
|---|---|
| `go_wp`, `fg_wp`, `punt_wp` | win probability of each choice |
| `go_wp_diff`, `fg_wp_diff`, `punt_wp_diff` | each against the best alternative |
| `go_boost` | how much going for it gains over the best kicking option |
| `first_down_prob` | conversion probability from the distribution |
| `wp_succeed`, `wp_fail` | WP after converting / failing |
| `fourth_down_recommendation` | the call |

```{r fourth-down, eval=FALSE}
library(cfbfastR)
library(dplyr)

pbp <- cfbfastR::load_cfb_pbp(2025)

pbp |>
  dplyr::filter(down == 4, !is.na(go_boost)) |>
  dplyr::select(pos_team, distance, yards_to_goal, go_wp, fg_wp, punt_wp,
                go_boost, first_down_prob, fourth_down_recommendation) |>
  dplyr::arrange(dplyr::desc(go_boost)) |>
  dplyr::slice_head(n = 20)
```

Sorting by `go_boost` finds the fourth downs where going for it was most clearly
right — and, sorted the other way, the punts that cost the most.

As with the two-point surface, **`go_boost` is more useful than the
recommendation**: a 0.001 edge and a 0.08 edge are different decisions, and only
the magnitude distinguishes them.

## QBR

A regression (`reg:squarederror`) on ten features, all of which are already EPA
components: `qbr_epa`, `sack_epa`, `pass_epa`, `rush_epa`, `pen_epa`, plus
`spread` and the one-hot era.

This is worth being clear about, because QBR is often described as a black box.
Here it is not a new measurement of anything — it is **a re-scaling of EPA
components onto the 0–100 scale people recognise**, fit to reproduce the
published QBR values. Everything it knows, it knows from EPA. If you want to
understand why a quarterback's QBR is what it is, look at his EPA splits; there
is nothing else in there.

## Everything degrades to `NA`

A design decision worth knowing about: none of these surfaces fail the pipeline.
If a model artifact cannot be fetched, or `arrow` is missing, or the punt table
is unavailable, or the game has no pre-game line, the columns are added as `NA`
rather than the stage erroring.

This means **you must check for `NA` rather than assuming a column is
populated** — particularly `vegas_wp` and the two decision surfaces, which need
a spread. The upside is that a network problem degrades your play-by-play instead
of destroying it.

```{r na-check, eval=FALSE}
pbp |>
  dplyr::summarise(dplyr::across(
    c(EPA, wpa, vegas_wp, cpoe, xpass, go_boost),
    ~ mean(is.na(.x))
  ))
```

## Seeing it applied

[Game on Paper](https://gameonpaper.com) surfaces most of this per game: the win
probability chart, per-play EPA, and fourth-down decision context. Its
[glossary](https://gameonpaper.com/glossary) defines the derived metrics in the
same terms used here.

## Frequently asked

**What is CPOE in college football?**
Completion percentage over expected — the gap between a passer's actual
completions and what the situation predicted, in percentage points. It separates
accuracy from difficulty of situation.

**What is pass rate over expected (`pass_oe`)?**
How much more often a team passes than its down, distance, score and clock
suggest. It measures aggression, not effectiveness.

**How accurate is the college field goal model?**
It gives about 81% at 34 yards and 54% at 50. It has no kicker, weather or
altitude input, and it flattens beyond roughly 57 yards where data runs out.

**Should I trust `fourth_down_recommendation`?**
Use `go_boost` instead. The recommendation has no margin, so it treats a
0.0001 edge as a decision.

## The series

1. [Expected points: how the model works](college-football-expected-points-model-fundamentals-part-i.html)
2. [How the model is built](college-football-expected-points-model-fundamentals-part-ii.html)
3. [A history of expected points models](college-football-expected-points-model-fundamentals-part-iii.html)
4. [What is EPA?](college-football-expected-points-model-fundamentals-part-iv.html)
5. [Win probability, WPA and `vegas_wp`](college-football-expected-points-model-fundamentals-part-v.html)
6. CPOE, xpass, field goals and fourth downs — this article

## Data and artifacts

- **Models** — [`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
- **Season play-by-play** — [`cfbfastR-data`](https://github.com/sportsdataverse/cfbfastR-data)
- **Source** — [`cfbfastR`](https://github.com/sportsdataverse/cfbfastR) · [`sportsdataverse-py`](https://github.com/sportsdataverse/sportsdataverse-py) · [`cfb4th`](https://github.com/sportsdataverse/cfb4th)
- **Applied, live** — [Game on Paper](https://gameonpaper.com)
- **Raw API** — [CollegeFootballData.com](https://collegefootballdata.com), courtesy of [@CFB_data](https://x.com/CFB_data)

## Citation

```
Gilani, S., Easwaran, A., Lee, J., and Hess, E. (2026). cfbfastR: Access College
Football Play by Play Data. R package version 3.0.0.9000.
https://cfbfastr.sportsdataverse.org
```

Authors, contributors and related SportsDataverse packages are listed on the
[package home page](../index.html).

```{r cleanup, include=FALSE}
options(old)
```
