---
title: "How the Expected Points Model Is Built (Part II)"
description: "Why expected points is a multiclass problem, why linear and binary logistic regression fail at it, and how the cfbfastR EP model went from 96 hand-built regression variables to eight gradient-boosted features."
author: "Saiem Gilani, Akshay Easwaran, Jared Lee"
date: "2026-09-03"
opengraph:
  image:
    src: "https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/social_card_cfbfastR_final_quote.png"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
old <- options(rmarkdown.html_vignette.check_title = FALSE)
```

[Part I](college-football-expected-points-model-fundamentals-part-i.html)
established what expected points is and showed the model's output. This article
is about *how you get there* — why the problem is shaped the way it is, why the
obvious approaches fail, and how the shipped model works.

The through-line: **predicting expected points is not a regression problem.** It
looks like one, because the answer is a number. It is not, and understanding why
explains every design decision that follows.

## The problem is multiclass, not continuous

The naive framing is "predict how many points the offense scores next." Try to
fit that directly and the problems appear immediately.

The target is not continuous. The next score in a half can only be one of seven
things, and their values are −7, −3, −2, 0, 2, 3 or 7. There is no such thing as
a 4.5-point outcome. A model that predicts 4.5 is not making a claim any
football game can satisfy — it is averaging over outcomes it should be
enumerating.

So the target is *categorical*, and expected points is a quantity you compute
**afterwards**, from the class probabilities:

$$\mathrm{EP} = \sum_{i=1}^{7} p_i \cdot v_i$$

That single reframing is the whole conceptual step. Everything below is about
estimating the $p_i$ well.

### Why not linear regression

Linear regression assumes the dependent variable is continuous and unbounded,
that the relationship between predictors and target is linear, that observations
are independent, and that predictors are not collinear.

Point values violate the first assumption outright. But even setting that aside,
predictions are unbounded — a linear model will happily tell you a situation is
worth 9.2 points, which no single scoring event can produce. And the
relationship is emphatically not linear: expected points changes far faster near
either goal line than it does at midfield.

### Why not binary logistic regression

Logistic regression fixes the boundedness problem. Model "does the offense score
a touchdown next, yes or no", and predictions live in $[0, 1]$ where
probabilities belong.

But we need seven probabilities that are *mutually exclusive and sum to one*.
Seven separately-fitted binary models give you seven numbers that do not sum to
anything in particular. You cannot form a valid expectation from them.

### Multinomial logistic regression

The classical fix is to fit the classes simultaneously against a reference
class, and pass the results through a softmax so they are normalised to sum to
one. Choose "No Score" as the reference, fit six sets of coefficients against
it, and the seventh falls out.

This is exactly what the 2020 model did — and what Yurko, Ventura and Horowitz
proposed for football in 2017 ([Part
III](college-football-expected-points-model-fundamentals-part-iii.html) covers
that lineage). It is a sound, interpretable and well-understood approach.

It is also where the 96 variables came from.

## The cost of linearity: 96 variables

A multinomial logit is linear in its predictors. It can only represent an
interaction if you build that interaction by hand, as a column. Football is
almost entirely interactions — 3rd and 2 is not "3rd down plus 2 yards to go",
it is its own situation — so the feature list grew to cover them:

| Group | Variables |
|---|---|
| Distance factors (yards to goal, log distance, goal-to-go, and their interaction) | 24 |
| Down, and down × distance, down × field position | 54 |
| Time factors, under-two-minute indicator, intercepts | 18 |
| **Total** | **96** |

Each bullet is six variables because six classes are fitted against the
reference. Every interaction anyone thought of had to be specified. Every
interaction nobody thought of was invisible to the model.

There were two further consequences worth naming:

- **The model could not see the score.** Game state was handled through
  observation *weights* — plays were weighted by score differential and by how
  many drives away the next score was — rather than as a feature.
- **Field goals needed a separate model**, a GAM on smoothed kick distance,
  because the main model had no way to represent "this is makeable."

## What changed: gradient boosting

The current model is an XGBoost `multi:softprob` classifier. It answers the same
seven-class question and produces probabilities the same way. What changes is
how the structure is found.

Boosted trees split on feature values, so an interaction is representable
without being specified: a tree that splits on `down_3`, then on `distance`,
then on `yards_to_goal` *is* a three-way interaction, discovered from the data.
The hand-built columns become unnecessary.

| | 2020 | 2026 |
|---|---|---|
| Estimator | `nnet::multinom` | XGBoost `multi:softprob` |
| Inputs | 96 constructed variables | **8 raw features** |
| Sees the score | no (weights only) | yes (`pos_score_diff_start`) |
| Training rows | 2014–2019, non-overtime | **2004–2025, 2,219,971 plays** |
| Field goals | separate GAM | separate gradient-boosted `fg_model` |

The eight features are simply:

```{r features, eval=FALSE}
c("TimeSecsRem", "yards_to_goal", "distance",
  "down_1", "down_2", "down_3", "down_4", "pos_score_diff_start")
```

Trained with `eta = 0.025`, `max_depth = 5` and 525 boosting rounds — which,
because there are seven classes, means the shipped booster contains **3,675
trees**. That is a useful sanity check when you load it:

```{r tree-count, eval=FALSE}
# xgb.dump() returns one element per NODE line, so length() is far larger than the
# tree count. Each tree is introduced by a `booster[i]` header -- count those.
sum(grepl("^booster\\[", xgboost::xgb.dump(ep_model)))   # 3675 = 525 rounds x 7 classes
```

### What the model actually leans on

Gain-based importance, computed from the shipped booster:

| Feature | Gain |
|---|---|
| `TimeSecsRem` | 30.9% |
| `yards_to_goal` | 21.9% |
| `down_4` | 14.5% |
| `down_3` | 9.4% |
| `down_1` | 7.9% |
| `down_2` | 6.9% |
| `pos_score_diff_start` | 5.0% |
| `distance` | 3.6% |

```{r importance, eval=FALSE}
imp <- xgboost::xgb.importance(model = ep_model)
imp[order(-imp$Gain), c("Feature", "Gain")]
```

Time is the largest single contributor, which is not obvious until you remember
what the target is: *the next score in this half*. With thirty minutes left
almost every drive has a successor; with thirty seconds left, "No Score" becomes
the overwhelming favourite regardless of field position. The model spends most
of its capacity on that.

Note also that `down_4` carries more gain than the other three downs combined
with each other — fourth down is where the possession itself is at risk, and the
distribution shifts hardest.

## A worked example, end to end

One situation, all the way through. First and 10 at your own 25, tied, thirty
minutes left in the half:

```{r worked, eval=FALSE}
x <- matrix(
  c(1800, 75, 10, 1, 0, 0, 0, 0), nrow = 1,
  dimnames = list(NULL, c("TimeSecsRem", "yards_to_goal", "distance",
                          "down_1", "down_2", "down_3", "down_4",
                          "pos_score_diff_start"))
)
# as.numeric(): predict() may return a 1 x 7 matrix here, and handing a matrix to
# data.frame() below would create seven columns against a seven-ROW outcome vector.
p <- as.numeric(predict(ep_model, x))
data.frame(
  outcome = c("TD", "Opp_TD", "FG", "Opp_FG", "Safety", "Opp_Safety", "No_Score"),
  prob    = round(p, 4),
  value   = c(7, -7, 3, -3, 2, -2, 0)
)
```

| Outcome | Probability | Value | Contribution |
|---|---|---|---|
| Touchdown | 0.3965 | +7 | +2.7753 |
| Opponent touchdown | 0.3182 | −7 | −2.2274 |
| Field goal | 0.1564 | +3 | +0.4691 |
| Opponent field goal | 0.1185 | −3 | −0.3555 |
| Safety | 0.0034 | +2 | +0.0068 |
| Opponent safety | 0.0024 | −2 | −0.0048 |
| No score | 0.0047 | 0 | 0.0000 |
| | | | **EP = +0.6635** |

Two thirds of a point, from a near-coin-flip between the two touchdown
outcomes. That is what "expected points" means concretely: not a prediction that
0.66 points will be scored, but the average over seven futures, six of which are
worth something.

### The score-margin feature, and a caveat

Because the current model sees the score, EP at a fixed spot varies with the
margin:

| Score margin | EP at own 25 |
|---|---|
| −21 | −1.38 |
| −7 | −0.18 |
| Tied | +0.66 |
| +7 | +1.71 |
| +21 | +2.71 |

**Read this carefully.** It does not mean leading causes you to score more from
your own 25. It means teams that are up 21 are, on average, better teams playing
worse opponents — the margin is partly a proxy for team quality, and the model
has no separate team-strength input to separate the two.

This is a real limitation and worth stating plainly: **EP is not opponent- or
quality-adjusted.** Any team-level metric built by summing EPA inherits that.
Adjusted versions, like the ones on
[Game on Paper](https://gameonpaper.com/year/2025/teams/offensive), apply
opponent and garbage-time corrections *on top of* raw EPA precisely because the
model itself does not.

## One artifact, two languages

The largest structural change is not statistical. The models are published once
as the
[`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
release and read by **both `cfbfastR` and `sportsdataverse-py`**
([#138](https://github.com/sportsdataverse/cfbfastR/issues/138)). The same play
gets the same EPA in R and in Python, and a retrain updates both by publishing,
without a release of either package.

Making that true required pinning down one thing that is easy to get wrong.
XGBoost emits classes in *its* training order, which is not the order `cfbfastR`
has always reported:

```{r class-order, eval=FALSE}
manifest$ep_class_contract$class_order
#> "TD" "Opp_TD" "FG" "Opp_FG" "Safety" "Opp_Safety" "No_Score"

manifest$ep_class_contract$cfbfastR_lev_order
#> "No_Score" "FG" "Opp_FG" "Opp_Safety" "Opp_TD" "Safety" "TD"

manifest$ep_class_contract$permutation_to_cfbfastR_lev_1based
#> 7 3 4 6 2 5 1
```

The permutation is **published in the manifest and read at runtime**, not
hard-coded. `cfbfastR` keeps a fallback copy, and a test asserts the two agree —
so if the bundle is ever retrained with a different class order, the build fails
loudly instead of silently reporting a touchdown probability as a safety
probability. There is no fixed point between the two orderings, which is to say
every position moves, which is to say a mistake here would be catastrophic and
completely invisible in the output.

If you score the booster yourself, apply the point values in the **bundle's**
order, as in the worked example above.

## Where the estimates are weakest

Being explicit about limits, since the point of this series is transparency:

- **No team-quality input.** See the score-margin caveat above.
- **PATs are treated as given.** The extra point is folded into the touchdown
  value of 7 rather than modelled.
- **Kickoffs assume a touchback baseline;** returns past the 25 are points added
  and shorter returns points lost.
- **Punts are not modelled separately.** Their effect is field position, which
  the EP model already handles well.
- **Overtime** is not represented in the next-score-in-half framing.
- **Era effects** are handled in the FG, two-point, QBR and fourth-down models
  through explicit `era0`–`era3` features, but the EP model itself has no era
  term — it pools 2004 through 2025.

## Next

[Part III](college-football-expected-points-model-fundamentals-part-iii.html)
traces where these ideas came from — Virgil Carter's 1970 paper, the field
position and down-distance models that followed, and the line that runs through
`nflscrapR` and `cfbscrapR` to the shared bundle described here.

## Data and artifacts

- **Models** — [`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
- **Season play-by-play** — [`cfbfastR-data`](https://github.com/sportsdataverse/cfbfastR-data)
- **Source** — [`cfbfastR`](https://github.com/sportsdataverse/cfbfastR) · [`sportsdataverse-py`](https://github.com/sportsdataverse/sportsdataverse-py)
- **Applied, live** — [Game on Paper](https://gameonpaper.com)
- **Raw API** — [CollegeFootballData.com](https://collegefootballdata.com), courtesy of [@CFB_data](https://x.com/CFB_data)

## Citation

```
Gilani, S., Easwaran, A., Lee, J., and Hess, E. (2026). cfbfastR: Access College
Football Play by Play Data. R package version 3.0.0.9000.
https://cfbfastr.sportsdataverse.org
```

Authors, contributors and related SportsDataverse packages are listed on the
[package home page](../index.html).

```{r cleanup, include=FALSE}
options(old)
```
