---
title: "Win Probability in College Football: wp, WPA and vegas_wp (Part V)"
description: "How the cfbfastR win probability model works, why it takes expected points as an input, what WPA measures that EPA cannot, and how the spread-aware vegas_wp differs from the naive wp."
author: "Saiem Gilani, Akshay Easwaran, Jared Lee"
date: "2026-09-03"
opengraph:
  image:
    src: "https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/social_card_cfbfastR_final_quote.png"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
old <- options(rmarkdown.html_vignette.check_title = FALSE)
```

**Win probability (WP) is the chance the team with the ball wins the game, given
the score, time, field position, down, distance and timeouts.** Where
[expected points](college-football-expected-points-model-fundamentals-part-i.html)
asks "how many points is this situation worth", win probability asks the question
that actually settles games — and the two can disagree sharply.

**WPA — win probability added — is the change in WP across a play**, exactly
parallel to
[EPA](college-football-expected-points-model-fundamentals-part-iv.html):

$$\mathrm{WPA} = \mathrm{WP}_{\text{after}} - \mathrm{WP}_{\text{before}}$$

## Why not just use EPA?

Because points and wins are the same thing only in the middle of a game.

Down 3 with a minute left, a 12-yard completion to the opponent's 30 is worth
modest EPA — a few tenths of a point. It is worth enormous WPA, because it moves
a field goal from unlikely to routine. Conversely, a 60-yard touchdown while up
28 in the fourth quarter is worth large EPA and almost exactly zero WPA. You had
already won.

The distinction matters whenever you rank *plays* or evaluate *decisions*. EPA
is the better measure of performance, because it is not distorted by score and
clock. WPA is the better measure of leverage, because it is.

## What the model uses

The naive model is a gradient-boosted binary classifier over twelve features:

| Feature | What it carries |
|---|---|
| `pos_score_diff_start` | current score margin |
| `ExpScoreDiff_Time_Ratio` | expected margin per second remaining |
| `adj_TimeSecsRem` | seconds left **in the game** |
| `TimeSecsRem` | seconds left in the half |
| `is_home` | home field |
| `pos_team_receives_2H_kickoff` | who gets the ball after halftime |
| `down`, `distance`, `yards_to_goal` | the situation |
| `pos_team_timeouts_rem_before`, `def_pos_team_timeouts_rem_before` | clock control |
| `period` | quarter |

Gain-based importance from the shipped booster:

| Feature | Gain |
|---|---|
| `pos_score_diff_start` | 53.1% |
| `ExpScoreDiff_Time_Ratio` | 25.6% |
| `is_home` | 11.5% |
| `pos_team_receives_2H_kickoff` | 3.5% |
| `yards_to_goal` | 2.0% |
| `adj_TimeSecsRem` | 1.5% |
| everything else | < 1% each |

Nearly 80% of the model is *score margin and score margin scaled by time*. That
is the correct shape for the problem — football win probability is overwhelmingly
about the lead and how long the opponent has to erase it — but it is worth
knowing before you over-interpret a WP swing on a 2nd-and-7.

### Expected points is an input to win probability

The second-most important feature is built from EP:

```{r expscorediff, eval=FALSE}
adj_TimeSecsRem          <- ifelse(half == 1, 1800 + TimeSecsRem, TimeSecsRem)
ExpScoreDiff             <- pos_score_diff_start + ep_before
ExpScoreDiff_Time_Ratio  <- ExpScoreDiff / (adj_TimeSecsRem + 1)
```

`ExpScoreDiff` is **the current margin plus what the current drive is worth** —
in effect, the score once this possession resolves. Dividing by time remaining
turns it into "how big is the effective lead, relative to how long is left to do
something about it."

So the four articles before this one are not a detour. The EP model feeds the WP
model directly: a better EP estimate is a better WP estimate.

## What the numbers look like

Scoring the naive model for a home team with all timeouts:

| Situation | WP |
|---|---|
| 1st & 10 own 25, tied, opening drive | **0.573** |
| 1st & 10 own 25, down 7, Q1 | 0.318 |
| 1st & 10 own 25, up 7, Q1 | 0.792 |
| 1st & 10 own 25, tied, Q4 5:00 | 0.648 |
| 1st & 10 own 25, down 3, Q4 2:00 | 0.255 |
| 4th & 2 opp 35, down 2, Q4 1:00 | 0.400 |
| 1st & goal opp 5, down 4, Q4 0:30 | 0.846 |
| 1st & 10 own 25, up 21, Q4 10:00 | 0.998 |

Two of these are worth pausing on.

**The opening drive is 0.573, not 0.500.** That is home-field advantage, and it
is why `is_home` carries 11.5% of the model's gain. A tied game is not a coin
flip if you are the home team.

**1st and goal from the 5, trailing by 4, with thirty seconds left is 0.846.**
Down by more than a field goal, needing a touchdown, with almost no time — and
you are a heavy favourite, because you are five yards from the win with four
downs to get there. No expected-points figure conveys that. This is precisely the
kind of situation where EPA and WPA part company.

```{r wp-columns, eval=FALSE}
library(cfbfastR)
library(dplyr)

pbp <- cfbfastR::load_cfb_pbp(2025)

pbp |>
  dplyr::filter(!is.na(wpa)) |>
  dplyr::select(pos_team, period, clock.minutes, pos_score_diff_start,
                play_text, wp_before, wp_after, wpa, vegas_wp) |>
  dplyr::arrange(dplyr::desc(abs(wpa))) |>
  dplyr::slice_head(n = 10)
```

Sorting by `abs(wpa)` is the quickest way to find the plays that decided a
season.

## WPA and its edge cases

`wp_after = wp_before + wpa`, and the defensive view is simply `1 - wp_after` —
win probability is zero-sum in a way expected points is not.

Two overlays are applied that a naive difference would get wrong:

- **Turnovers.** Possession changes hands, so the raw model output is for the
  *other* team. WPA is computed against the complement, the same way EPA negates
  expected points across a change of possession.
- **End of half and end of period.** A play that ends a half does not hand the
  next situation to anyone in the usual sense; `wpa_half_end` handles the
  transition so the last snap of a half is not credited with a phantom swing.

These are the same two places EPA needs care, for the same reason.

## `vegas_wp`: what the market knew

The naive model treats both teams as anonymous. Kickoff between the best team in
the country and the worst is 50/50 plus home field, because nothing in those
twelve features says who is playing.

That is a defensible choice — it measures *in-game* leverage without prejudice —
but it is wrong as a forecast. So a second model adds one feature:

```{r spread-time, eval=FALSE}
elapsed_share <- pmax((3600 - adj_TimeSecsRem) / 3600, 0)
spread_time   <- pos_team_spread * exp(-4 * elapsed_share)
```

`spread_time` is the pre-game line, **exponentially decayed as the game runs
down**. The market's opinion should dominate before any football has been
played, and count for less once sixty minutes of evidence exist. This is a
faithful port of the `nflfastR` / `sportsdataverse-py` construction, decay
constant included.

| Game point | `adj_TimeSecsRem` | Elapsed | `spread_time` (7-point line) |
|---|---|---|---|
| Kickoff | 3600 | 0.00 | 7.00 |
| End of Q1 | 2700 | 0.25 | 2.58 |
| Halftime | 1800 | 0.50 | 0.95 |
| End of Q3 | 900 | 0.75 | 0.35 |
| 2:00 left | 120 | 0.97 | 0.15 |
| Final gun | 0 | 1.00 | 0.13 |

### The feature decays much faster than its influence

It is tempting to read that table as "by halftime the spread counts for 13% of
what it did at kickoff." **That is wrong, and it is worth understanding why.**

`spread_time` is an *input*. What matters is the model's response to it, which is
a learned, non-linear function — and a tree ensemble is free to react sharply to
small values. Measuring the actual effect, as the gap between `vegas_wp` and the
naive `wp` in an otherwise identical state, expressed as a share of the same gap
at kickoff:

| Line | Kickoff | Q1 end | Halftime | Q3 end | Q4 10:00 |
|---|---|---|---|---|---|
| +7 | 100% | 61% | 62% | 27% | −15% |
| +10 | 100% | 70% | 55% | 28% | 16% |
| +14 | 100% | 84% | 64% | 36% | 23% |
| +21 | 100% | 85% | 67% | 43% | 31% |
| −14 (underdog has ball) | 100% | 99% | 82% | 61% | 50% |

**Roughly 55–80% of the spread's influence survives to halftime**, not 13%. The
model compensates for the shrinking feature by responding to it more steeply. The
exponential decay sets the *scale* the model trains against; it does not set the
weight the model gives it.

### Where this gets unreliable

The same table shows a real limitation, and it is not one to paper over.

The decay is **not monotonic**, and at small lines late in games it **changes
sign**: a 7-point favourite is given *lower* `vegas_wp` than the naive model
through the fourth quarter (−15% at 10:00, −20% at 5:00). A 3-point line is worse
still — the effect is unstable from the end of the first quarter onward.

This is tree-model behaviour in a thin region: by the fourth quarter
`|spread_time|` for a modest line has decayed to a few tenths of a point, and the
model has little training signal to separate 0.2 from −0.2. The artifact is
largest exactly where the feature is smallest.

Practical guidance:

- `vegas_wp` is most trustworthy **early**, and for **large** lines — which is
  also when it adds the most over the naive model.
- Late in close games with a small spread, prefer `wp`. The scoreboard has told
  you everything the line was going to, and `vegas_wp` is reading noise.
- Do not difference `vegas_wp` and `wp` to "extract" the market's contribution.
  That quantity is not stable enough to interpret.

This is the same split `nflfastR` draws between `wp` and `vegas_wp`, and
`cfbfastR` keeps `wp_before` **completely unchanged** — `vegas_wp` is added
alongside rather than replacing it, with `vegas_wpa` and `vegas_wp_after`
derived using the same turnover and half-end overlays.

### Two sign conventions that must agree

This is the sort of thing that produces a confidently backwards win probability
chart:

- CFBD's `spread` is **negative when the home team is favoured**.
- The model reads **positive `spread_time` as the team in possession being
  favoured**.

Those are not the same convention, and the conversion depends on who has the
ball. The mapping was verified across 83 games rather than assumed — worth
knowing if you build `spread_time` yourself.

### When is `vegas_wp` `NA`?

Wherever the game has no pre-game line: the ESPN play-by-play path, and CFBD
before 2013. The column is added as `NA` rather than the stage failing, so
filter on it rather than assuming it is populated.

```{r which-wp, eval=FALSE}
# in-game leverage, team-agnostic -- use for play and decision evaluation
pbp$wp_before

# forecast that knows who is playing -- use for "should they have won?"
pbp$vegas_wp
```

**Which should you use?** For evaluating a decision or a play's leverage, the
naive `wp` is usually right: it asks what this situation is worth to anybody. For
"how surprising was this result", `vegas_wp` is right, because a 14-point
comeback against a 21-point favourite is a different event from the same comeback
against an even opponent.

## Limits

- **Neither model knows anything about the teams beyond the spread.** The naive
  model knows nothing at all; `vegas_wp` knows only the market's pre-game number.
- **No injury, weather, or personnel information** enters either.
- **Overtime is not modelled** in the same framing.
- **The tails are the hardest part.** A model trained on 2.2 million plays sees
  very few genuine 0.99 → 0.20 collapses, so extreme values are the least
  well-estimated. Treat 0.997 and 0.999 as "almost certainly won", not as a
  meaningful ordering.
- **Timeouts are in the model but carry under 1% of gain**, which is probably an
  underweight relative to how much late-game coaches agonise over them.

## Seeing it applied

Every game page on [Game on Paper](https://gameonpaper.com) renders the win
probability chart for that game, with the play-by-play beneath it showing the WP
swing on each snap — the plays with the largest `|wpa|` are the ones the chart's
cliffs correspond to.

## Frequently asked

**What is a good WPA for a single play?**
Anything above ±0.10 is a genuinely big play; ±0.30 is a game-turning one.
Regular-season games are usually decided by a handful of plays above 0.15.

**Why is win probability not 50% at kickoff?**
Home-field advantage. The naive model puts the home team near 0.57 in a tied
game on the opening drive. With `vegas_wp`, the spread moves it further.

**What is the difference between `wp` and `vegas_wp`?**
`wp` ignores who is playing; `vegas_wp` adds the pre-game spread, decayed as the
game progresses. `vegas_wp` is `NA` when no line is available.

**Can win probability be wrong?**
It is a probability, not a prediction. A team at 0.95 loses one time in twenty,
and over a season that will happen many times. The model is wrong only if teams
at 0.95 fail to win about 95% of the time.

## Next

Part VI covers the derived surfaces built on these two models — completion
probability and CPOE, expected pass rate (`xpass` / `pass_oe`), the field goal
and two-point models, fourth-down decisions, and QBR.

## Data and artifacts

- **Models** — [`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts) (`wp_naive.ubj`, `wp_spread.ubj`)
- **Season play-by-play** — [`cfbfastR-data`](https://github.com/sportsdataverse/cfbfastR-data)
- **Source** — [`cfbfastR`](https://github.com/sportsdataverse/cfbfastR) · [`sportsdataverse-py`](https://github.com/sportsdataverse/sportsdataverse-py)
- **Applied, live** — [Game on Paper](https://gameonpaper.com)
- **Raw API** — [CollegeFootballData.com](https://collegefootballdata.com), courtesy of [@CFB_data](https://x.com/CFB_data)

## Citation

```
Gilani, S., Easwaran, A., Lee, J., and Hess, E. (2026). cfbfastR: Access College
Football Play by Play Data. R package version 3.0.0.9000.
https://cfbfastr.sportsdataverse.org
```

Authors, contributors and related SportsDataverse packages are listed on the
[package home page](../index.html).

```{r cleanup, include=FALSE}
options(old)
```
