---
title: "What Is EPA in College Football? Expected Points Added Explained (Part IV)"
description: "EPA (expected points added) is the change in expected points across a single play. How it is calculated, what a good EPA per play looks like, the sign conventions that trip people up, and how to compute it yourself with cfbfastR."
author: "Saiem Gilani, Akshay Easwaran, Jared Lee"
date: "2026-09-03"
opengraph:
  image:
    src: "https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/social_card_cfbfastR_final_quote.png"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
old <- options(rmarkdown.html_vignette.check_title = FALSE)
```

**EPA — expected points added — is the change in expected points across a single
play.** If a team's situation was worth 0.66 expected points before the snap and
1.92 after it, the play was worth **+1.26 EPA**. It is the standard way to put a
points value on one football play, and nearly every modern team and player
metric is built by summing or averaging it.

$$\mathrm{EPA} = \mathrm{EP}_{\text{after}} - \mathrm{EP}_{\text{before}}$$

[Part I](college-football-expected-points-model-fundamentals-part-i.html) covers
what expected points is and
[Part II](college-football-expected-points-model-fundamentals-part-ii.html) how
the model estimates it. This article is about turning that into per-play value.

## Why EPA instead of yards

Yards treat every situation as identical. EPA does not, and the difference is
the entire point.

A 3-yard gain on 3rd and 2 continues a drive. A 3-yard gain on 3rd and 8 ends
one. Both are "3 yards". Their EPA differs by more than two points, because the
model knows what happens next in each case.

Start from a routine situation — 1st and 10 at your own 25, tied, thirty minutes
left in the half — worth **+0.66** expected points. Here is what various outcomes
of that single play are worth:

| Play outcome | Resulting situation | EP after | **EPA** |
|---|---|---|---|
| Gain of 40 | 1st & 10, own 65 | +3.57 | **+2.90** |
| Gain of 25 | 1st & 10, midfield | +2.61 | **+1.95** |
| Gain of 15 | 1st & 10, own 40 | +1.92 | **+1.25** |
| Gain of 8 | 2nd & 2, own 33 | +1.62 | **+0.95** |
| Gain of 10 | 1st & 10, own 35 | +1.55 | **+0.89** |
| Gain of 5 | 2nd & 5, own 30 | +0.96 | **+0.30** |
| Gain of 3 | 2nd & 7, own 28 | +0.30 | **−0.36** |
| Incompletion | 2nd & 10, own 25 | −0.02 | **−0.69** |
| Sack for −7 | 2nd & 17, own 18 | −1.13 | **−1.79** |
| Interception at the spot | Opponent 1st & 10 | −0.66 | **−1.33** |
| Pick six risk: INT returned to your 10 | Opponent 1st & 10, your 10 | −4.96 | **−5.62** |

Look at rows four and five. **A gain of 8 yards is worth more EPA than a gain of
10.** Eight yards leaves 2nd and 2; ten yards leaves 1st and 10 five yards
further downfield. The model prefers 2nd and 2, and it is right to — a 2nd and 2
converts at a far higher rate than a fresh set of downs needing ten.

That is not a quirk. It is the clearest possible demonstration that **down and
distance carry more value than raw field position** in the middle of the field,
and no yards-based statistic can express it.

Notice also that a 3-yard gain on 1st and 10 is *negative*. Gaining yards is not
automatically good. Gaining fewer than the situation demands is a loss.

## The third-down case

The same idea, sharper. 3rd and 3 at the opponent's 35 is worth **+2.75**:

| Outcome | Resulting situation | EP after | **EPA** |
|---|---|---|---|
| Convert, gain of 5 | 1st & 10, opponent 30 | +3.98 | **+1.23** |
| Stuffed for no gain | 4th & 3, opponent 35 | +0.88 | **−1.87** |
| Turnover on downs | Opponent ball, their 35 | −1.55 | **−4.30** |

A three-yard gain here is worth +1.23. A no-gain is worth −1.87. **The same
three yards that were worth −0.36 on 1st and 10 are worth +1.23 on 3rd and 3.**

## Sign conventions

This is where most confusion lives, so it is worth being explicit.

**EPA is always signed from the perspective of the team with the ball at the
snap.** Positive EPA is good for that offense, negative is bad.

Consequences that catch people out:

- **Defensive EPA is the same number negated.** A defense that allows −0.15
  EPA/play is *good*. When building defensive metrics you either flip the sign
  or say "EPA allowed" and remember lower is better. `cfbfastR` reports the
  offense-signed value; the flip is yours to make.
- **On a turnover, expected points for the new offense is negated** before
  differencing. That is why an interception at the spot costs −1.33 rather than
  −0.66: you lose your own +0.66 *and* hand the opponent an equivalent
  situation.
- **Return men differ by play type.** On a kickoff the returning team is the
  possession team, so returner EPA is positive-good. On a punt return the
  punting team is the possession team at the snap, so the same raw column needs
  negating for the returner. This is the single most common error in
  player-level EPA aggregation.
- **Penalties** are folded into the play they attach to, so a defensive holding
  on an incompletion produces one net EPA, not two.

## EPA, EPA per play, and total EPA

Three different statistics that get called "EPA":

| Statistic | What it measures | Use it for |
|---|---|---|
| **EPA** (one play) | value of a single play | play-level analysis |
| **Total EPA** | sum over plays | volume — season value, workhorse backs |
| **EPA per play** | mean over plays | efficiency — comparing teams and players fairly |

A quarterback with 1,200 total EPA over 500 dropbacks (2.4 per play) is not
comparable to one with 1,200 over 900 dropbacks. Per-play is almost always the
right comparison; total EPA is the right way to credit workload.

## What is a good EPA per play?

For **teams**, from the 2025 FBS season as ranked on
[Game on Paper](https://gameonpaper.com/year/2025/teams/offensive) (adjusted
offensive EPA per play, 135 teams):

| Rank | Team | Adj. EPA/play |
|---|---|---|
| 1 | Vanderbilt | +0.35 |
| 4 | Ohio State | +0.26 |
| 11 | Miami | +0.22 |
| 21 | Georgia | +0.15 |
| 51 | Pittsburgh | +0.04 |
| 81 | Kennesaw State | −0.01 |
| 135 | Ball State | −0.31 |

So, as a rule of thumb for team offenses:

- **above +0.25** — elite, top-five
- **+0.10 to +0.25** — very good
- **around 0.00** — average; the median FBS offense sits near zero by construction
- **below −0.20** — among the worst in the country

The full range across 135 FBS teams is roughly **−0.31 to +0.35**. That is a
narrower spread than newcomers expect: the gap between the best and worst
offense in the country is about two thirds of a point per snap.

Note these are *adjusted* — corrected for opponent strength, home field and
garbage time. Raw EPA per play runs wider and flatters teams with easy
schedules, which is exactly why the adjustment exists.

## Computing it yourself

`cfbfastR` ships EPA on every play, so normally you just read it:

```{r epa-columns, eval=FALSE}
library(cfbfastR)
library(dplyr)

pbp <- cfbfastR::load_cfb_pbp(2025)

pbp |>
  dplyr::filter(!is.na(EPA)) |>
  dplyr::select(pos_team, def_pos_team, down, distance, yards_to_goal,
                play_text, ep_before, ep_after, EPA) |>
  dplyr::slice_head(n = 10)
```

Team efficiency, offense and defense, from raw play data:

```{r team-epa, eval=FALSE}
offense <- pbp |>
  dplyr::filter(!is.na(EPA)) |>
  dplyr::group_by(team = pos_team) |>
  dplyr::summarise(plays = dplyr::n(),
                   epa_per_play = mean(EPA),
                   total_epa = sum(EPA), .groups = "drop")

defense <- pbp |>
  dplyr::filter(!is.na(EPA)) |>
  dplyr::group_by(team = def_pos_team) |>
  dplyr::summarise(def_epa_per_play = mean(EPA), .groups = "drop")
  # lower is better -- these are offense-signed

offense |>
  dplyr::inner_join(defense, by = "team") |>
  dplyr::mutate(net = epa_per_play - def_epa_per_play) |>
  dplyr::arrange(dplyr::desc(net)) |>
  dplyr::filter(plays > 300)
```

Note the `plays > 300` filter. Per-play averages over small samples are noisy;
a receiver with four targets and +2.0 EPA per target has told you nothing.

And reproducing the worked table above, straight from the published model:

```{r reproduce, eval=FALSE}
# ep_at() as defined in Part I
before <- ep_at(75, down = 1, dist = 10)          # 1st & 10, own 25

data.frame(
  outcome = c("gain 8", "gain 10", "gain 15"),
  ep_after = c(ep_at(67, down = 2, dist = 2),
               ep_at(65, down = 1, dist = 10),
               ep_at(60, down = 1, dist = 10))
) |> transform(EPA = ep_after - before)
```

## What EPA does not do

Stated plainly, because a metric is only useful if you know its edges:

- **It is not opponent-adjusted.** Raw EPA credits a touchdown against a bad
  defence exactly like one against a good defence.
- **It has no team-quality input**, so score margin partly proxies for it — see
  the caveat in
  [Part II](college-football-expected-points-model-fundamentals-part-ii.html).
- **It does not assign credit within a play.** A 40-yard completion produces one
  EPA figure; how much belongs to the quarterback, the receiver, the line or the
  coverage bust is a separate modelling problem.
- **It is descriptive, not predictive.** High EPA per play describes what
  happened. Predicting next week needs stability adjustments, because
  explosiveness regresses harder than efficiency.
- **Garbage time distorts it** unless you filter or weight for win probability.

## Seeing it applied

[Game on Paper](https://gameonpaper.com) computes all of this for every FBS game.
Any game page shows EPA per play, success rate and a win probability chart, with
a play-by-play listing `ep_before`, `ep_after` and EPA for every snap — the
table at the top of this article, for a real game. Season leaderboards rank
every team by [offensive](https://gameonpaper.com/year/2025/teams/offensive) and
[defensive](https://gameonpaper.com/year/2025/teams/defensive) EPA per play, and
players by [passing](https://gameonpaper.com/year/2025/players/passing),
[rushing](https://gameonpaper.com/year/2025/players/rushing) and
[receiving](https://gameonpaper.com/year/2025/players/receiving) EPA per play.

## Frequently asked

**Is a high EPA good?**
For an offense, yes — positive EPA means the play improved its scoring
expectation. For a defense the sign flips: you want the offense's EPA against
you to be negative.

**What is a good EPA per play?**
For an FBS offense, above +0.25 is elite and around 0.00 is average. The full
2025 range was roughly −0.31 to +0.35.

**What is the difference between EPA and EPA per play?**
EPA is one play's value; total EPA sums it (a volume measure); EPA per play
averages it (an efficiency measure).

**Can EPA be negative on a play that gained yards?**
Yes, and often. A 3-yard gain on 1st and 10 is worth −0.36 EPA, because it
leaves a worse down and distance than the offense started with.

**Does EPA account for the opponent?**
Not on its own. Adjusted versions correct for opponent, home field and garbage
time on top of raw EPA.

## Next

Part V covers **win probability** — the same modelling idea applied to the
outcome that actually matters, why `vegas_wp` exists alongside `wp`, and how
WPA differs from EPA in late-game situations.

## Data and artifacts

- **Models** — [`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
- **Season play-by-play** — [`cfbfastR-data`](https://github.com/sportsdataverse/cfbfastR-data)
- **Source** — [`cfbfastR`](https://github.com/sportsdataverse/cfbfastR) · [`sportsdataverse-py`](https://github.com/sportsdataverse/sportsdataverse-py)
- **Applied, live** — [Game on Paper](https://gameonpaper.com)
- **Raw API** — [CollegeFootballData.com](https://collegefootballdata.com), courtesy of [@CFB_data](https://x.com/CFB_data)

## Citation

```
Gilani, S., Easwaran, A., Lee, J., and Hess, E. (2026). cfbfastR: Access College
Football Play by Play Data. R package version 3.0.0.9000.
https://cfbfastr.sportsdataverse.org
```

Authors, contributors and related SportsDataverse packages are listed on the
[package home page](../index.html).

```{r cleanup, include=FALSE}
options(old)
```
