---
title: "CPOE, xpass, Field Goals and Fourth Downs: The Derived Models (Part VI)"
description: "The six models cfbfastR builds on top of expected points and win probability: completion probability and CPOE, expected pass rate, field goals, two-point conversions, fourth-down decisions and QBR — with their real outputs and their limits."
author: "Saiem Gilani, Akshay Easwaran, Jared Lee"
date: "2026-09-03"
opengraph:
  image:
    src: "https://raw.githubusercontent.com/sportsdataverse/cfbfastR-data/main/themes/social_card_cfbfastR_final_quote.png"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
old <- options(rmarkdown.html_vignette.check_title = FALSE)
```

[Expected points](college-football-expected-points-model-fundamentals-part-i.html)
and [win probability](college-football-expected-points-model-fundamentals-part-v.html)
are the foundation. Everything in this article is built on top of them, and all
of it ships in the same
[`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
bundle.

There is a useful distinction to draw first. Some of these are **models** — they
predict a thing and you read the number. Others are **analytic surfaces**: they
build the game state that *would* follow a choice, score it with the WP model,
and compare. Fourth down and the two-point decision are surfaces, and they are
ports of [`cfb4th`](https://github.com/sportsdataverse/cfb4th).

## The bundle at a glance

| Artifact | Objective | Features | Produces |
|---|---|---|---|
| `ep_model` | `multi:softprob` (7) | 8 | `ep_before`, `ep_after`, `EPA` |
| `wp_naive` | `binary:logistic` | 12 | `wp_before`, `wpa` |
| `wp_spread` | `binary:logistic` | 13 | `vegas_wp` |
| `cfb_cp_model` | `binary:logistic` | 8 | `cp`, `cpoe` |
| `xpass_model` | `binary:logistic` | 7 | `xpass`, `pass_oe` |
| `fg_model` | `binary:logistic` | 5 | field-goal probability |
| `two_pt_model` | `binary:logistic` | 4 | `prob_2pt` |
| `fd_model` | `multi:softprob` (**76**) | 9 | fourth-down gain distribution |
| `qbr_model` | `reg:squarederror` | 10 | QBR |

Plus `punt_distribution.parquet`, which is not a model at all — see below.

## Completion probability and CPOE

`cp` is the probability a given pass attempt is completed, from down, distance,
field position, score, clock, home field, period and a passing-down indicator.
**CPOE** is the difference between what happened and what was expected, on a
percentage-point scale:

$$\mathrm{CPOE} = 100 \times (\text{completion} - \mathrm{cp})$$

| Situation | `cp` |
|---|---|
| 1st & 10, own 25 | 0.639 |
| 3rd & 3, opponent 40 | 0.595 |
| 2nd & 20, own 20 | 0.627 |
| 3rd & 12, own 30 | 0.489 |

CPOE is the standard way to separate a quarterback's accuracy from his
situation. A 65% completion rate throwing 3rd-and-12s is a much better
performance than 65% throwing 1st-and-10s, and CPOE is what says so.

Note that the model has **no receiver, pressure, or air-yards input** — it is a
situational baseline, not a throw-difficulty model. A quarterback checking down
constantly will post good CPOE.

## Expected pass rate (`xpass` and `pass_oe`)

`xpass` is the probability a play *is* a pass, given the situation — a model of
tendency, not of quality. `pass_oe` (pass rate over expected) is
`100 * (pass - xpass)`, and measures how much more pass-happy a team is than its
situations warrant.

| Situation | `xpass` |
|---|---|
| 1st & goal, opponent 3 | 0.141 |
| 2nd & 2, own 33 | 0.247 |
| 3rd & 2, opponent 40 | 0.255 |
| 1st & 10, own 25 | 0.373 |
| 2nd & 8, own 27 | 0.536 |
| 3rd & 8, opponent 40 | 0.782 |

The spread is the point: 14% on 1st and goal from the 3, 78% on 3rd and 8. Any
raw pass-rate statistic is mostly measuring which of these situations a team
found itself in. `pass_oe` removes that.

### Two era encodings, one set of cutpoints

The bundle carries the rule era two ways. `xpass` and the two-point model take an
**ordinal** `era` column, 0–3; the FG and QBR models take **one-hot** `era0`–`era3`
dummies.

The cutpoints are the same for all of them — a single shared constant,
`ERA_BOUNDS = (2006, 2013, 2020)`, giving 2004–2006, 2007–2013, 2014–2020 and
2021 on. Both encodings derive from it, so there is no divergence to trip over,
only two shapes of the same fact.

If you score these models by hand, match each one's encoding: an ordinal `3`
where four dummy columns are expected does not fail loudly. `cfbfastR` handles it
internally.

## The field goal model

Five features: `yards_to_goal` and the one-hot era. Kick distance is
`yards_to_goal + 17`.

| `yards_to_goal` | Kick | P(make) |
|---|---|---|
| 2 | 19 | 0.956 |
| 10 | 27 | 0.893 |
| 17 | 34 | 0.809 |
| 20 | 37 | 0.754 |
| 25 | 42 | 0.675 |
| 30 | 47 | 0.614 |
| 33 | 50 | 0.542 |
| 38 | 55 | 0.455 |
| 43 | 60 | 0.366 |
| 48 | 65 | 0.366 |

A 50-yarder is a coin flip, which is a useful number to carry around.

**Era matters, and it is not a rounding effect.** Kicking has genuinely improved,
and the model captures it — the same 42-yard attempt across the four eras:

| Era | P(make), 42 yards |
|---|---|
| `era0` (pre-2006) | 0.556 |
| `era1` (2006–2012) | 0.598 |
| `era2` (2013–2019) | 0.612 |
| `era3` (2020+) | 0.675 |

Twelve percentage points from the oldest era to the current one. The era features
carry about 20% of the model's gain between them, so scoring a historical kick
with `era3` — or a modern one with `era0` — is a real error, not a nicety.

**The tail runs out at 57 yards.** From `yards_to_goal` 40 outward the model
returns a constant **0.3656** — a 57-yard and a 70-yard attempt get identical
probabilities. Nobody attempts many 65-yarders except in desperation, so there is
nothing to learn from and the model extrapolates flat. Treat anything past a
57-yard kick as "the model has run out of evidence" rather than an estimate.

**Two small monotonicity violations.** Across 58 steps the curve rises twice
where it should fall: a 49-yard kick is rated 0.5882 against 0.5647 for a
48-yarder, and 51 yards beats 50. Both are sub-3-point artifacts of tree
splitting rather than anything meaningful, but if you are differencing adjacent
yard lines — as the fourth-down surface does when it weighs a make against a miss
— be aware the curve is not strictly monotone.

The model also has **no kicker identity, weather or altitude input**. A kick in
Laramie and a kick in a dome get the same number.

## The two-point model

Four features: `posteam_spread`, `posteam_total`, `pos_score_diff`, `era`. It
produces `prob_2pt`, which feeds the two-point decision surface —
`two_pt_wp`, `xp_wp`, `two_pt_wp_diff` and `two_pt_recommendation`.

**`posteam_total` is not the game total.** It is the possessing team's *implied
team total*, built from the line and the over/under:

```{r posteam-total, eval=FALSE}
home_total <- (homeTeamSpread + overUnder) / 2
away_total <- (overUnder - homeTeamSpread) / 2
```

A 55-point game with a 7-point spread gives roughly 24 and 31, so the feature
lives around **14–42**, not 45–65. This matters if you score the model yourself:
feed it a game total and every prediction lands off the end of the training
support, where the model is flat and meaningless.

Over the real support it is the model's strongest input:

| `posteam_total` | `prob_2pt` |
|---|---|
| 14 | 0.4521 |
| 20 | 0.5024 |
| 24 | 0.5213 |
| 30 | 0.5100 |
| 32 | 0.5774 |
| 38 | 0.5780 |

Across a realistic grid — spread ±21, team total 16–40, score margin ±10, all
eras — the model runs **0.365 to 0.592**, mean **0.511**. Its `base_score` is
**0.4816**, the observed success rate in training, so it sits about **3
percentage points above its own base rate**, which is what conditioning should
do. Split counts back that up: `posteam_total` 54, `pos_score_diff` 32,
`posteam_spread` 20.

### The one real gap: `era` is never used

`era` is in the feature list and in none of the 40 trees:

```{r two-pt-splits, eval=FALSE}
xgboost::xgb.importance(model = two_pt_model)$Feature
#> "posteam_total"  "pos_score_diff"  "posteam_spread"     # no "era"
```

Sweeping era 0 → 3 with everything else fixed moves the prediction by **0.0000**.
Two-point conversion rates have moved across 2004–2025 and this model cannot
express that. With `max_depth = 2`, `min_child_weight = 40` and 40 rounds over a
relatively small set of attempts, the other three features win every split.

**Practical consequence.** The surface's arithmetic is sound — verified
**bit-identical to `sportsdataverse-py` to eight decimal places** — but `cfb4th`'s
rule has no margin, and will recommend two on a difference of 0.0001. **Read
`two_pt_wp_diff` and treat anything under a couple of percentage points as a coin
flip**, rather than the bare recommendation.

## The fourth-down decision

This is the most elaborate surface in the package, and the 76-class model is the
interesting part.

`fd_model` is a `multi:softprob` classifier over **76 classes: yards gained, from
−10 to +65**. It does not predict whether a fourth down is converted. It predicts
the *whole distribution* of what happens if you go for it.

The go branch then:

1. Scores the 76-class distribution for the fourth-down state.
2. Expands the play into **one hypothetical game state per outcome** — every
   yardage result gets its own down, distance and field position.
3. Scores each with the win probability model.
4. Averages, weighted by the distribution.

That average is `go_wp` — the win probability of going for it, integrated over
everything that could happen.

The punt branch is different again. **`cfb4th` does not model punts.** It joins
an empirical distribution of punt end-yardlines from
`punt_distribution.parquet`, expands to one state per landing spot, scores each,
and averages. The field-goal branch uses the FG model above, weighting a make
against a miss and the field position each produces.

The outputs:

| Column | Meaning |
|---|---|
| `go_wp`, `fg_wp`, `punt_wp` | win probability of each choice |
| `go_wp_diff`, `fg_wp_diff`, `punt_wp_diff` | each against the best alternative |
| `go_boost` | how much going for it gains over the best kicking option |
| `first_down_prob` | conversion probability from the distribution |
| `wp_succeed`, `wp_fail` | WP after converting / failing |
| `fourth_down_recommendation` | the call |

```{r fourth-down, eval=FALSE}
library(cfbfastR)
library(dplyr)

pbp <- cfbfastR::load_cfb_pbp(2025)

pbp |>
  dplyr::filter(down == 4, !is.na(go_boost)) |>
  dplyr::select(pos_team, distance, yards_to_goal, go_wp, fg_wp, punt_wp,
                go_boost, first_down_prob, fourth_down_recommendation) |>
  dplyr::arrange(dplyr::desc(go_boost)) |>
  dplyr::slice_head(n = 20)
```

Sorting by `go_boost` finds the fourth downs where going for it was most clearly
right — and, sorted the other way, the punts that cost the most.

As with the two-point surface, **`go_boost` is more useful than the
recommendation**: a 0.001 edge and a 0.08 edge are different decisions, and only
the magnitude distinguishes them.

## QBR

A regression (`reg:squarederror`) on ten features, all of which are already EPA
components: `qbr_epa`, `sack_epa`, `pass_epa`, `rush_epa`, `pen_epa`, plus
`spread` and the one-hot era.

This is worth being clear about, because QBR is often described as a black box.
Here it is not a new measurement of anything — it is **a re-scaling of EPA
components onto the 0–100 scale people recognise**, fit to reproduce the
published QBR values. Everything it knows, it knows from EPA. If you want to
understand why a quarterback's QBR is what it is, look at his EPA splits; there
is nothing else in there.

## Everything degrades to `NA`

A design decision worth knowing about: none of these surfaces fail the pipeline.
If a model artifact cannot be fetched, or `arrow` is missing, or the punt table
is unavailable, or the game has no pre-game line, the columns are added as `NA`
rather than the stage erroring.

This means **you must check for `NA` rather than assuming a column is
populated** — particularly `vegas_wp` and the two decision surfaces, which need
a spread. The upside is that a network problem degrades your play-by-play instead
of destroying it.

```{r na-check, eval=FALSE}
pbp |>
  dplyr::summarise(dplyr::across(
    c(EPA, wpa, vegas_wp, cpoe, xpass, go_boost),
    ~ mean(is.na(.x))
  ))
```

## Seeing it applied

[Game on Paper](https://gameonpaper.com) surfaces most of this per game: the win
probability chart, per-play EPA, and fourth-down decision context. Its
[glossary](https://gameonpaper.com/glossary) defines the derived metrics in the
same terms used here.

## Frequently asked

**What is CPOE in college football?**
Completion percentage over expected — the gap between a passer's actual
completions and what the situation predicted, in percentage points. It separates
accuracy from difficulty of situation.

**What is pass rate over expected (`pass_oe`)?**
How much more often a team passes than its down, distance, score and clock
suggest. It measures aggression, not effectiveness.

**How accurate is the college field goal model?**
In the modern era (`era3`, 2020+) it gives about 81% at 34 yards and 54% at 50.
Older eras are materially lower — 42 yards is 0.68 in `era3` against 0.56 in
`era0`. It has no kicker, weather or altitude input, and flattens beyond roughly
57 yards where data runs out.

**Should I trust `two_pt_recommendation`?**
Not on its own — `cfb4th`'s rule has no margin and will recommend two on a
difference of 0.0001. Read `two_pt_wp_diff` and treat small margins as coin
flips. The underlying `prob_2pt` model is reasonable (mean 0.511 against a 0.4816
base rate) but ignores `era`.

**Should I trust `fourth_down_recommendation`?**
Use `go_boost` instead. The recommendation has no margin, so it treats a
0.0001 edge as a decision.

## The series

1. [Expected points: how the model works](college-football-expected-points-model-fundamentals-part-i.html)
2. [How the model is built](college-football-expected-points-model-fundamentals-part-ii.html)
3. [A history of expected points models](college-football-expected-points-model-fundamentals-part-iii.html)
4. [What is EPA?](college-football-expected-points-model-fundamentals-part-iv.html)
5. [Win probability, WPA and `vegas_wp`](college-football-expected-points-model-fundamentals-part-v.html)
6. CPOE, xpass, field goals and fourth downs — this article

## Data and artifacts

- **Models** — [`cfb_model_artifacts`](https://github.com/sportsdataverse/sportsdataverse-data/releases/tag/cfb_model_artifacts)
- **Season play-by-play** — [`cfbfastR-data`](https://github.com/sportsdataverse/cfbfastR-data)
- **Source** — [`cfbfastR`](https://github.com/sportsdataverse/cfbfastR) · [`sportsdataverse-py`](https://github.com/sportsdataverse/sportsdataverse-py) · [`cfb4th`](https://github.com/sportsdataverse/cfb4th)
- **Applied, live** — [Game on Paper](https://gameonpaper.com)
- **Raw API** — [CollegeFootballData.com](https://collegefootballdata.com), courtesy of [@CFB_data](https://x.com/CFB_data)

## Citation

```
Gilani, S., Easwaran, A., Lee, J., and Hess, E. (2026). cfbfastR: Access College
Football Play by Play Data. R package version 3.0.0.9000.
https://cfbfastr.sportsdataverse.org
```

Authors, contributors and related SportsDataverse packages are listed on the
[package home page](../index.html).

```{r cleanup, include=FALSE}
options(old)
```
