---
title: "Creating Fourth Down Tendency Plots Using cfbfastR"
description: "A rundown of the Big XII returning coaches fourth down tendencies"
author: "Michael Egle <br><a href='https://twitter.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&label=%40deceptivespeed_&logo=x&style=for-the-badge' alt='@deceptivespeed_' /></a> <a href='https://github.com/michaelegle' target='blank'><img src='https://img.shields.io/github/followers/michaelegle?color=eee&logo=Github&style=for-the-badge' alt='@michaelegle' /></a>"
date: "9/5/2020"
opengraph:
  image: 
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@deceptivespeed_"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hey everyone, my name is Michael and over the summer I worked on a daily series of plots using ```ggplot``` and the ```cfbfastR``` package. One of my favorite plots I put together was the fourth down tendency plot for various head coaches. This visualization was inspired by Michael Lopez doing the same thing for NFL coaches. This tutorial is going to walk through how they're put together. If you haven't already, you should read the introduction tutorial that Parker made to get used to the data and download the package.

## First, we'll have to install and import the necessary packages
```{r install_packages, message = FALSE, warning = FALSE}
if (!requireNamespace('pacman', quietly = TRUE)){
  install.packages('pacman')
}
pacman::p_load(tidyverse, cfbfastR)
```


We are going to load in data for seasons 2014-2020, it'll take between 45-90 seconds to run.
```{r load_2014_2020, warning = FALSE}
tictoc::tic()
pbp <- data.frame()
seasons <- 2014:2020
progressr::with_progress({

  pbp <- cfbfastR::load_cfb_pbp(seasons)
})
tictoc::toc()
```

Next, we'll need to get the coaching information, so we'll use the ```cfbd_coaches``` function:
```{r fourth_down_coaches}
coaches <- purrr::map_dfr(seasons,function(x){cfbfastR::cfbd_coaches(year = x)})

```
`cfbd_coaches()` already returns each coach's `conference`, so no extra lookup is needed.

```{r fourth_down_coaches2}
coaches <- coaches |>
  dplyr::mutate(coach = paste(first_name, last_name, sep = " ")) |>
  dplyr::filter(games >= 6) |>
  dplyr::select(coach, school, year, conference)
```

Since interim coaches are included in the coaches dataframe, we'll set the cutoff at coaching 6 or more games.

We only have a couple more steps to make our graph. Next, we need to add our coaches to the ```pbp``` dataframe using join functions
```{r fourth_down_join}
pbp <- pbp |>
  dplyr::inner_join(coaches, by = c("offense_play" = "school", "year" = "year"))
```

Now we can filter down to only fourth down plays, then we'll add columns to determine if the play was a punt, FGA, or the team went for it.
```{r fourth_down_mutate}
down4 <- pbp |>
  dplyr::filter(down == 4) |>
  dplyr::mutate(fga = ifelse(str_detect(play_type, "Field Goal"),
                             1, 0),
                punt = ifelse(play_type == "Punt", 1, 0),
                attempt = ifelse(rush == 1 | pass == 1, 1, 0),
                play = dplyr::case_when(fga == 1 ~ "FG Attempt",
                                 punt == 1 ~ "Punt",
                                 attempt == 1 ~ "Go"))
```

And now we've got all the info we need to make the graph! The code below is listed for current Big 12 head coaches, but you can alter it to show whichever coaches you want.
```{r fourth_down_plot, fig.alt="Big 12 Coaches's Fourth Down Tendencies | CFP Era"}
down4 |>
  dplyr::filter(!is.na(play)) |>
  dplyr::filter(coach %in% c("Matt Campbell", "Steve Sarkisian", "Lincoln Riley", "Chris Klieman", "Matt Wells",
                     "Neal Brown", "Les Miles", "Mike Gundy", "Gary Patterson", "Dave Aranda")) |>
  dplyr::filter(distance <= 5, distance > 0) |>
  ggplot(aes(x = distance, y = 100 - yards_to_goal, color = play)) +
  geom_jitter() +
  facet_wrap(. ~ coach) +
  theme_bw() +
  labs(x = "Yards to Go",
       title = "Big 12 Coaches's Fourth Down Tendencies | CFP Era",
       subtitle = "Data from @cfbfastR",
       caption = "Visualization by Michael Egle (@deceptivespeed_)",
       color = "Decision") +
  scale_y_continuous(labels = c("Own 20", "Own 40", "Opp 40",
                               "Opp 20", "Endzone"),
                     breaks = c(20, 40, 60, 80, 100)) +
  theme(axis.title.y = element_blank())
```

Looks good! Small sample size but can easily be built upon. Hopefully you found this tutorial helpful and can make some more cool CFB related visualizations with ```ggplot2```

# **Our Authors**
-   [Saiem Gilani](https://x.com/saiemgilani)
<a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a>
<a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>
-   [Akshay Easwaran](https://x.com/akeaswaran)
<a href='https://x.com/akeaswaran' target='blank'><img src='https://img.shields.io/twitter/follow/akeaswaran?color=blue&amp;label=%40akeaswaran&amp;logo=x&amp;style=for-the-badge' alt='@akeaswaran'/></a>
<a href='https://github.com/akeaswaran' target='blank'><img src='https://img.shields.io/github/followers/akeaswaran?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@akeaswaran'/></a>
-   [Jared Lee](https://x.com/JaredDLee)
<a href='https://x.com/JaredDLee' target='blank'><img src='https://img.shields.io/twitter/follow/JaredDLee?color=blue&amp;label=%40JaredDLee&amp;logo=x&amp;style=for-the-badge' alt='@JaredDLee'/></a>
<a href='https://github.com/JaredDLee' target='blank'><img src='https://img.shields.io/github/followers/JaredDLee?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@JaredDLee'/></a>
-   [Eric Hess](https://x.com/arbitanalytics)
<a href='https://x.com/arbitanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/arbitanalytics?color=blue&amp;label=%40arbitanalytics&amp;logo=x&amp;style=for-the-badge' alt='@arbitanalytics'/></a>
<a href='https://github.com/arbitanalytics' target='blank'><img src='https://img.shields.io/github/followers/arbitanalytics?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@arbitanalytics'/></a>

## **Our Contributors**
-   [Michael Egle](https://x.com/deceptivespeed_)
<a href='https://x.com/deceptivespeed_' target='blank'><img src='https://img.shields.io/twitter/follow/deceptivespeed_?color=blue&amp;label=%40deceptivespeed_&amp;logo=x&amp;style=for-the-badge' alt='@deceptivespeed_'/></a>
-   [Nate Manzo](https://x.com/cfbnate)
<a href='https://x.com/cfbnate' target='blank'><img src='https://img.shields.io/twitter/follow/cfbnate?color=blue&amp;label=%40cfbnate&amp;logo=x&amp;style=for-the-badge' alt='@cfbnate'/></a>
-   [Jason DeLoach](https://x.com/CFBNumbers)
<a href='https://x.com/CFBNumbers' target='blank'><img src='https://img.shields.io/twitter/follow/CFBNumbers?color=blue&amp;label=%40CFBNumbers&amp;logo=x&amp;style=for-the-badge' alt='@CFBNumbers'/></a>
-   [Tej Seth](https://x.com/tejfbanalytics)
<a href='https://x.com/tejfbanalytics' target='blank'><img src='https://img.shields.io/twitter/follow/tejfbanalytics?color=blue&amp;label=%40tejfbanalytics&amp;logo=x&amp;style=for-the-badge' alt='@tejfbanalytics'/></a>
-   [Conor McQuiston](https://x.com/ConorMcQ5)
<a href='https://x.com/ConorMcQ5' target='blank'><img src='https://img.shields.io/twitter/follow/ConorMcQ5?color=blue&amp;label=%40ConorMcQ5&amp;logo=x&amp;style=for-the-badge' alt='@ConorMcQ5'/></a>
-   [Tan Ho](https://x.com/_TanHo)
<a href='https://x.com/_TanHo' target='blank'><img src='https://img.shields.io/twitter/follow/_TanHo?color=blue&amp;label=%40_TanHo&amp;logo=x&amp;style=for-the-badge' alt='@_TanHo'/></a>
-   [Keegan Abdoo](https://x.com/KeeganAbdoo)
<a href='https://x.com/KeeganAbdoo' target='blank'><img src='https://img.shields.io/twitter/follow/KeeganAbdoo?color=blue&amp;label=%40KeeganAbdoo&amp;logo=x&amp;style=for-the-badge' alt='@KeeganAbdoo'/></a>
-   [Matt Spencer](https://x.com/Maatspencer)
<a href='https://x.com/Maatspencer' target='blank'><img src='https://img.shields.io/twitter/follow/Maatspencer?color=blue&amp;label=%40Maatspencer&amp;logo=x&amp;style=for-the-badge' alt='@Maatspencer'/></a>
-   [Sebastian Carl](https://x.com/mrcaseb)
<a href='https://x.com/mrcaseb' target='blank'><img src='https://img.shields.io/twitter/follow/mrcaseb?color=blue&amp;label=%40mrcaseb&amp;logo=x&amp;style=for-the-badge' alt='@mrcaseb'/></a>
-   [John Edwards](https://x.com/John_B_Edwards)
<a href='https://x.com/John_B_Edwards' target='blank'><img src='https://img.shields.io/twitter/follow/John_B_Edwards?color=blue&amp;label=%40John_B_Edwards&amp;logo=x&amp;style=for-the-badge' alt='@John_B_Edwards'/></a>
-   [Brad Hill](https://x.com/bradisblogging)
<a href='https://x.com/bradisblogging' target='blank'><img src='https://img.shields.io/twitter/follow/bradisblogging?color=blue&amp;label=%40bradisblogging&amp;logo=x&amp;style=for-the-badge' alt='@bradisblogging'/></a>

## **Citation**

To cite the [**`cfbfastR`**](https://cfbfastR.sportsdataverse.org/) R package in publications, use:

BibTex Citation

```bibtex
@misc{cfbfastr,
  author = {Saiem Gilani, Akshay Easwaran, Jared Lee, Eric Hess},
  title = {cfbfastR: The SportsDataverse's R Package for College Football Data},
  url = {https://cfbfastR.sportsdataverse.org/},
  year = {2026}
}
```

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) - college football
-   [**hoopR**](https://hoopR.sportsdataverse.org/) - men's basketball
-   [**wehoop**](https://wehoop.sportsdataverse.org/) - women's basketball
-   [**baseballr**](https://baseballr.sportsdataverse.org/) - baseball
-   [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) - hockey
-   [**oddsapiR**](https://oddsapiR.sportsdataverse.org/) - betting odds
-   [**sportyR**](https://sportyR.sportsdataverse.org/) - playing surfaces
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) - the Python package
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) - the R meta-package
