---
title: "Introduction to cfbfastR"
description: "Getting started with using <code>cfbfastR</code> and college football analytics."
author: "Saiem Gilani <br><a href='https://twitter.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=twitter&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image: 
    src: "https://github.com/saiemgilani/cfbfastR-data/blob/master/themes/social_card_cfbfastR_final_quote.png?raw=true"
  twitter:
    site: "@cfbfastR"
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load_current_gh("r-lib/pkgapi")
library(pkgapi)
pkg <- pkgapi::map_package(path = "../")
library(dplyr)
exported <- pkg$defs %>% dplyr::filter(exported == TRUE)
cfbd_funcs <- sum(stringr::str_detect(exported$file,"cfbd")) - sum(stringr::str_detect(exported$file,"cfbd_pbp_data"))
pkg_name <- "saiemgilani/cfbfastR"
url <- paste0("https://raw.githubusercontent.com/", pkg_name, "/master/DESCRIPTION")


x <- readLines(url)
remote_version <- gsub("Version:\\s*", "", x[grep('Version:', x)])
```

Hey folks,

Welcome to the football analytics community! I'm Saiem Gilani, one of the [authors](https://saiemgilani.github.io/cfbfastR/authors.html "Authors and contributors to cfbfastR") of [`cfbfastR`](https://twitter.com/cfbfastR "Follow @cfbfastR on Twitter! Tag us in your tweets using cfbfastR and we'll share it!"), and I hope to give the community a high-quality resource for accessing college football data for statistical analysis, football research, and more. I am excited to show you some of what you can do with this edition of the package.

### **Installing R and RStudio**
  1. Head to [https://cran.r-project.org](https://cran.r-project.org "The Comprehensive R Archive Network")
  2. Select the appropriate link for your operating system (Windows, Mac OS X, or Linux)
    - **Windows** - Select base and download the most recent version
    - **Mac OS X** - Select Latest Release, but check to make sure your OS is the correct version. Look  through Binaries for Legacy OS X Systems if you are on an older release
    - **Linux** - Select the appropriate distro and follow the installation instructions
  3. Head to [RStudio.com](https://www.rstudio.com/products/rstudio/download/#download "Download the appropriate version of RStudio (Free) for your operating system to use with R")
  4. Follow the associated download and installation instructions for RStudio.
  5. Start peering over the [RStudio IDE Cheatsheet]("https://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf"). *An IDE is an integrated development environment.*
  
#### **Load and install the necessary packages**

```{r load_cfbfastR_tidy}
if (!requireNamespace('pacman', quietly = TRUE)){
  install.packages('pacman')
}
pacman::p_load_current_gh("saiemgilani/cfbfastR")
pacman::p_load(tidyverse, zoo, ggimage, gt)
```

### **The Data**
There are generally speaking **three** college football data sources accessed from this package:

-   [`cfbfastR-data` repo](https://github.com/saiemgilani/cfbfastR-data) [![Twitter Follow](https://img.shields.io/twitter/follow/cfbfastR?color=blue&label=%40cfbfastR&logo=twitter&style=for-the-badge)](https://twitter.com/cfbfastR)
-   [College Football Data API](https://collegefootballdata.com) [![Twitter Follow](https://img.shields.io/twitter/follow/cfb_data?color=blue&label=%40cfb_data&logo=twitter&style=for-the-badge)](https://twitter.com/cfb_data)
-   [ESPN](https://espn.com)

#### **Function names indicate the data source**

* Functions that use the [`cfbfastR-data` repository](https://github.com/saiemgilani/cfbfastR-data) will contain `_cfb` or `cfb_` in the function name and would be considered loading functions for the play-by-play data.

* Functions that use the CFB Data API start with `cfbd_` by convention and should be assumed as `get` functions. 

* Functions that use one of ESPN's APIs start with `espn_` by convention and should be assumed as `get` functions. There are only two of these functions so far: `espn_ratings_fpi()` and `espn_metrics_wp()`

However, there is only one data *provider* involved for most game data, ESPN's data provider.

As of `cfbfastR` version `r remote_version`, the package exports `r nrow(exported)` functions. The bulk (\~`r cfbd_funcs`) of the functions within the package serve as the unofficial R API client for the [College Football Data API](https://collegefootballdata.com).

#### **CFB Data now requires an API key (it's free)**
-   Since April 1, 2021, the College Football Data API requires [key authentication](https://collegefootballdata.com/key), but the **key is free to acquire and use**.

-   [Follow the instructions](https://collegefootballdata.com/key) and wait for your API key to be delivered to the e-mail account associated with your key.

#### **Using the CFB Data key**

You can save the key for consistent usage by adding `CFBD_API_KEY=XXXX-YOUR-API-KEY-HERE-XXXXX` to your .REnviron file (easily accessed via [**`usethis::edit_r_environ()`**](https://usethis.r-lib.org/reference/edit.html)).

Run [**`usethis::edit_r_environ()`**](https://usethis.r-lib.org/reference/edit.html) and THEN paste the following in the new script that pops up (with**out** quotations)

``` {r, eval = FALSE}
CFBD_API_KEY = XXXX-YOUR-API-KEY-HERE-XXXXX
```

For less consistent usage, save your API key as the environment variable `CFBD_API_KEY` (with quotations) at the beginning of every session, using a command like the following.

```{r, eval = FALSE}
Sys.setenv(CFBD_API_KEY = "XXXX-YOUR-API-KEY-HERE-XXXXX")
```


If you have ever worked with the now archived [`cfbscrapR`](https://github.com/saiemgilani/cfbscrapR) package, most of the functions in ```cfbfastR``` should be fairly familiar with some slight changes.


We are going to load in data for seasons 2014-2020, it'll take between 45-90 seconds to run.
```{r load_2014_2020, warning = FALSE}
tictoc::tic()
pbp <- data.frame()
seasons <- 2014:2020
progressr::with_progress({
future::plan("multisession")
  pbp <- cfbfastR::load_cfb_pbp(seasons)
})
tictoc::toc()
```
